Specific Sensitivity to Rare and Extreme Events: Quasi-Complete Black Swan Avoidance vs Partial Jackpot Seeking in Rat Decision-Making

Version 1

Abstract

Abstract

Most studies assessing animal decision-making under risk rely on probabilities that are typically larger than 10%. To study Decision-Making in uncertain conditions, we explore a novel experimental and modelling approach that aims at measuring the extent to which rats are sensitive - and how they respond - to outcomes that are both rare (probabilities smaller than 1%) and extreme in their consequences (deviations larger than 10 times the standard error). In a four-armed bandit task, stochastic gains (sugar pellets) and losses (time-out punishments) are such that extremely large - but rare - outcomes materialize or not depending on the chosen options. All rats feature both limited diversification, mixing two options out of four, and sensitivity to rare and extreme outcomes despite their infrequent occurrence, by combining options with avoidance of extreme losses (Black Swans) and exposure to extreme gains (Jackpots). Notably, this sensitivity turns out to be one-sided for the main phenotype in our sample: it features a quasi-complete avoidance of Black Swans, so as to escape extreme losses almost completely, which contrasts with an exposure to Jackpots that is partial only. The flip side of observed choices is that they entail smaller gains and larger losses in the frequent domain compared to alternatives. We have introduced sensitivity to Black Swans and Jackpots in a new class of augmented Reinforcement Learning models and we have estimated their parameters using observed choices and outcomes for each rat. Adding such specific sensitivity results in a good fit of the selected model - and simulated behaviors that are close - to behavioral observations, whereas a standard Q-Learning model without sensitivity is rejected for almost all rats. This model reproducing the main phenotype suggests that frequent outcomes are treated separately from rare and extreme ones through different weights in Decision-Making.

Introduction

Although exploration and exploitation are key drivers of observed behavior when humans and other animals interact repeatedly with their environment (see e.g. [26] for an overview), not much is arguably known about how they are affected by outcomes that are both very rare and very consequential, that is, Rare and Extreme Events (REE thereafter). While risk is a central dimension in the related and enormous literature that contributes to social sciences and neurosciences, as it should, most studies on Decision-Making rely on lotteries for which the frequencies of outcomes are typically larger than 10% (e.g. [9], [10], [37], [49]; see also [19] for exceptions when choices are, however, not repeated). This is admittedly a major limitation in view of the fact that many living species experience events that are not only much rarer but that also entail potentially very large -, negative as well as positive impacts. The subsequent evolutionary consequences for - and possibly the extinction of large groups or species constitute an important example (see e.g. [18]). On the brighter side, human evolution has been punctuated by the design and spread of artefacts, a few of which have allowed big leaps forward on the way to material comfort (see e.g. [27]).

To the extent that both their consequences and their frequencies are highly uncertain, because historical data either lacks or is scarce, such REE are challenging for Decision-Making. Too little is known about whether animal species, including humans, are sensitive to - and how they cope with - such extremely infrequent and consequential events when making choices. In this study, we focus on rats, which have been formerly used to assess Decision-Making processes in various behavioral tasks (e.g. [6], [1], [45], [47], [23], [30], [43], [20], [54]; see also [22] on the relevance of animal models for human Decision-Making), but mostly because they allow further neurobiological manipulations in future studies. We have designed a novel experiment in which rats interact repeatedly with their laboratory environment through a 4-armed bandit task. In this task, they face REE that occur with a frequency smaller than 1% and, equally importantly, that are associated with extreme outcomes in terms of both gains and losses - deviations that are larger than 10 times the standard error. Following the design of the Iowa Gambling Task for human participants (see [4]), related tasks for rodents typically assume that the less frequent events have probability values that are about or larger than 10% (see e.g. [1], [34] and [53]).

In addition to dealing with low-probability and high-consequence events, the key feature of our experimental design is that it allows to test whether, in such a context, sensitivity to REE favors choosing frequently exposures to uncertainty that limit the consequences of extremely “negative” risks - or losses - while leaving open the possibilities to benefit from “positive” and large risks - or gains (called “Anti-fragile” exposures in [48]). Symmetrically, therefore, an option that rules out extreme gains but may trigger very large losses (labeled Black Swans) is labeled Fragile in our design, and it would be chosen frequently or exclusively when sensitivity to REE is rather low or nonexistent. More precisely, complete exposure to Jackpots and complete avoidance of Black Swans both define the Anti-fragile option, while the reverse configuration is representative of the Fragile one (i.e. complete avoidance of Jackpots and complete exposure to Black Swans, rather oddly). We also consider Robust and Vulnerable options: the former, which may seem to correspond to resilience, rules out exposure to any REE while the latter exposes to both REE, positive and negative. As such, our approach is of interest in an evolutionary perspective, which would ask how often the Anti-fragile option is indeed selected by animals and humans, and whether there are natural foundations for Anti-fragile choices that possibly relate to fitness and are embodied in the brain.

Importantly, our experimental paradigm distinguishes three domains as follows: “normal events” (NE), which occur with approximately 90% probability, “rare events” (RE) happening with about 10% probability, and REE which occur with a probability smaller than 1%. Each option is defined by a relationship between the magnitudes of rewards and the frequencies at which they occur. In, say, the gain domain, while one of the two options dominates the other for NE, this dominance weakens when RE are added to to picture and is eventually reversed if REE occur. To put it simply, the option that is dominated for NE and RE but turns out to dominate when REE occur is such that the less frequent the reward, the more consequential it is in terms of larger values: we therefore label this option convex. It follows that convex options expose to Jackpots and, by the same token, avoids Blacks Swans while, symmetrically, concave options avoids Jackpots and expose to Black Swans. Therefore, the Anti-fragile option is convex because it combines accelerating and possibly large gains (labeled Jackpots thereafter), together with decelerating and limited losses, when probabilities decrease - that is, when the number of trials needed to observe such gain occurrences increase. Here, to repeat, convexity does not refer to the shape of utility or other value functions, but refers to the fact that the magnitude of gains/losses, is accelerating/decelerating against decreasing probability (i.e. increasing number of trials). In sum, our experimental design is based on the assumptions that: 1) with perfect information about the probability distributions of all exposures, value-maximizing participants whose preferences are represented by any non-decreasing value function, would choose concave gains in the domain restricted to NE, due to first-order stochastic dominance. In other words, in the NE domain, it is more interesting to choose the concave options; 2) they would continue to do so in the RE domain for any non-decreasing and concave value function, due to second-order Stochastic Dominance; 3) however, in the full domain with REE, value-maximizing subjects endowed with any non-decreasing and concave value function are predicted to choose convex gains, because of a reversal in second-order Stochastic Dominance. In other words, exposure to Jackpots and avoidance of Black Swans is by design costly and disadvantageous when REE do not materialize.

Because of our specific interest in REE on both positive (rewards) and negative (losses) values, our framing integrates gains (sugar pellets) and losses (time-out punishment preventing the rats to earn pellets). One attractive feature of our design is that it delivers two direct and fairly natural measures that help interpreting behavioral data. First, Total Sensitivity to REE measures the extent to which rats take into account REE by combining convex options with avoidance of extreme lossees (Black Swans) and exposure to extreme gains (Jackpots). Formally, Total Sensitivity adds up the fractions of convex choices in terms of both gains and losses. Second, One-sided Sensitivity to REE captures choices that preferably, and asymmetrically, favor either the seeking of large gains (Jackpot Seeking) or the avoidance of large losses (Black Swan Avoidance). That is, One-sided Sensitivity is defined as the difference between the fraction of convex choices in gains and that in losses. While the Anti-fragile option has by design maximal Total Sensitivity and the Fragile option has zero Total Sensitivity, they both exhibit zero One-Sided Sensitivity. In contrast, the other alternatives, i.e. the Robust and Vulnerable options, have both middle Total Sensitivity but differ in that they respectively exhibit REE avoidance (negative One-Sided Sensitivity) and REE seeking (positive One-Sided Sensitivity). Of course, rats have the possibility to combine all four options over the course of the experiments.

Using our setup on a sample of 20 rats, with about 6000 stimuli per rat over 41 sessions, we document two sets of behavioral results. First, all rats diversify their choices but in a limited fashion across the possible set of options, primarily 2 out of the 4 available. Most rats (19 out of 20) exhibit moderate to high levels of Total Sensitivity. This means that most rats choose to diversify their choices across options that often combine some exposure to extreme gains and some avoidance of extreme losses. More strikingly, most rats (13 out of 20) tend to exhibit quasi-complete Black Swan avoidance mixed with partial Jackpot seeking: they have high Total Sensitivity to REE and favor avoiding almost certainly Black Swans even though this entails less frequent exposure to Jackpots (negative One-Sided Sensitivity to REE), by mixing mostly the Anti-fragile and Robust options. We interpret such a behavior as evidence of the differential treatment of rare and extreme gains and losses. Consistent with such an interpretation, we also find that Total Sensitivity and Black Swan avoidance averaged over all sessions are reinforced after rare and extreme losses are effectively experienced in the trials.

We next examine the behavioral observations though the lens of Reinforcement Learning models, which are theoretical benchmarks in the exploration-exploitation literature (see e.g. [46]). More specifically, the modelling contribution of this study is to introduce sensitivity to Black Swans and Jackpots in a new class of augmented Q-Learning models and we estimate their parameters using observed choices and outcomes for each rat. Interestingly, the model selected through information criteria has a distinctive feature: it separates NE/RE from REE in the decision to pick one option rather than any of the other. In other words, the selected model says that most rats behave as if they attach different weights to normal/rare outcomes and to rare and extreme outcomes, especially losses, in Decision-Making.

Results

Measures of sensitivities to REE in a four-armed bandit task

Figure 1 summarizes the main features of our experimental setting. In panel (a) is depicted a schematic representation of the Skinner box, in which all rats performed all 41 experimental sessions . Each session consisted of about 120 trials. Each rat had the possibility at all times to poke into 4 holes, each corresponding to a different sequence of gains and losses, drawn at random for each rat from a fixed distribution across sessions. Gains were sugar pellets while losses were time-out punishments (time period during which nose-poke remained inactive, thus entailing an opportunity cost in terms of pellets not consumed) measured in seconds (see the Material and Methods sections on Experimental Model and Subject Details, Experimental Method Details). In the following we use “options” for possible choices and “choices” for observed choices. The key feature of the four possible options associated to each available hole was that they were combinations of convex (that is, accelerating) and/or concave (that is, decelerating) gains and losses with decreasing frequencies of occurrence, in panel (b) of Figure 1 (see the Material and Methods section on Modelling and Statistical Analysis for the Jensen gap as a quantitative measure of convexity). While, as shown in the first and last columns, values for losses and gains differed for convex and concave options, the associated probabilities were assumed to be identical. They were set, as seen in the middle column, to {p1 = 0.25 − ε/3, p2 = 0.2 − ε/3, p3 = 0.05 − ε/3, ε} for both concave and convex gains. Parameter ε is the (very small) ex-ante probability of the largest outcome that we label the Jackpot for the 80-pellet gain, and the Black Swan for the 240-second loss. Both outcomes are denoted Rare and Extreme Events (REE in short) (as a convenient shortcut, therefore, the Black Swan is thought of as a rare and extreme loss, to differentiate it from a rare and extreme gain).

Experimental design - panel (a), schematic representation of the rat behavioral operant chamber equipped with four holes, each associated with a specific exposure to gains and losses (Anti-fragile, Robust, Vulnerable and Fragile); panel (b), values and frequencies for convex (green)/concave (red) gains/losses, with the largest probabilities highlighted in blue for Normal Events (20 to 25%, NE domain), Rare Events in yellow (5%, RE domain) and Rare and Extreme Events in pink (ε-probability smaller than 1%, REE domain); panel (c), representation and label of the four possible options, with increasing gains/losses (up/down) against decreasing probabilities (right/left): “Robust” has convex losses (avoidance of negative REE called Black Swan) and concave gains (avoidance of positive REE called Jackpot), “Anti-fragile” has convex losses and gains (Black Swan avoidance and Jackpot seeking), “Fragile” has concave losses and gains (Black Swan seeking and Jackpot avoidance) and “Vulnerable” has concave losses (Black Swan seeking) and convex gains (Jackpot seeking); panel (d), representation of two measures of Sensitivity to REE, i.e. “Total Sensitivity” (y-axis) and “One-sided Sensitivity” (x-axis) and examples, where green dot represents sensitivities associated with an hypothetical near-Anti-fragile profile (close to full Jackpot seeking and complete Black Swan avoidance), red dot represents sensitivities associated with an hypothetical near-Fragile profile (close to full Black Swan seeking and Jackpot avoidance), and blue dot represents sensitivities associated with an hypothetical choice mixing mostly Anti-fragile and Robust (near complete Black Swan avoidance and partial Jackpot seeking) - see the Material and Methods section on Modelling and Statistical Analysis for details

The next largest values (that is, 12 pellets and 15 seconds, vs 5 pellets and 36 seconds) are labeled a Rare Event (RE), since their total probability is around 10% and the associated values are moderate compared to REE. Note that for concave gains and convex losses the RE and REE values coincide. This is to contrast with the Jackpot and Black Swan that occur only if convex gains and concave losses are chosen, respectively. With total probability about 90%, therefore, the remaining set of lowest values in the first two rows (highlighted in blue in panel (b) of Figure 1) for concave and convex options are labeled Normal Events (NE), and these are the typical events most considered in the literature. Combining gains and losses that are either convex or concave then delivers 4 options.

The 4 options that correspond to the 4 holes in the Skinner box are pictured in panel (c) of Figure 1, with the following meaning. On the horizontal x-axis are reported (in decreasing order moving away from the origin) the ex-ante probabilities that are unknown to the subject, which only observes the outcome that is measured on the vertical y-axis (In contrast with [10], we do not give rats any cue about the events’ frequencies during the training sessions, which are set by us as experimenters. This makes our setting ecologically closer to experience-based designs.). For convenience, gains appear in the upper-right quadrant while losses appear in the lower-left one. For example, the option named Anti-fragile (after [48]) at the top of panel (c) is composed of convex gains and convex losses: the right part depicts the points in the (x, y) axis are {p1, 1}, {p2, 3}, {p3, 12}, {ε, 80}, corresponding to the fourth column in panel (b); the left part corresponds to the points {ε, −15}, {p3, −15}, {p2, −12}, {p3, −6}, where “negative” values are interpreted as losses in seconds. Looking at the green curves that summarize the Anti-fragile option, then, one visualizes convexity directly from the properties that gains accelerate moving right from the origin, while losses decelerate moving left from the origin, with decreasing probability. Alternatively, decreasing probability can be thought of as increasing numbers of trials that are needed to experience the associated gain or loss. As a consequence, the Anti-fragile option potentially exposes to the Jackpot but avoids exposure to the Black Swan (While in panel (c) convex (concave) gains and losses are identified by color, in green (red), a geometric description might be useful as well. For example the property that a monotone, upward or downward, convex (concave) curve lies below (above) its chords joining any pair of points can be used to identify green (red) curves.).

Symmetrically, the Fragile option in the lower part of panel (c) in Figure 1 has both gains and losses that are concave (in red). Gains are {p1, 2}, {p2, 4}, {p3, 5}, {ε, 5}, which means that gains plateau at 5 pellets, compared with 80 for the Jackpot attached to convex gains in the Anti-fragile option. In the loss domain, the Fragile option delivers {ε, −240}, {p3, −36}, {p2, −9}, {p3, −3}: it exposes to the Black Swan with an awfully long delay of 240 seconds. For the Fragile option, then, gains decelerate while losses accelerate, with decreasing probability, so that it potentially exposes to the Black Swan but avoids exposure to the Jackpot. Finally, the ways the Robust and Vulnerable options combine convex/concave gains and losses may be viewed as symmetric: the Robust choice protects from the Black Swan but at the same time misses the opportunity to get the jackpot, while the Vulnerable option potentially delivers both REE since both gains and losses accelerate with decreasing probability (or increasing number of trials).

In sum, the four options in panel (c) of Figure 1 are therefore combinations of the convex/concave gains/losses in panel (b): the Anti-fragile option at the top is convex both in gains and in losses, while the Robust option (left) is convex only in the loss domain. On the other hand, the Vulnerable exposure (right) has convex gains only, while the Fragile option (bottom) is concave both in gains and in losses. This implies that the Jackpot of 80 pellets may materialize only when either the Anti-fragile or the Vulnerable exposure are picked, while the Black Swan of 240 seconds may be experienced only by rats choosing either the Fragile or the Vulnerable exposures. In the Material and Methods section on Modelling and Statistical Analysis, it is shown that convex gains/losses dominate concave gains/losses in the sense of Stochastic Dominance, provided that the probability of REE ε is larger than about 0.3%. In our design, the values for ε we assumed to meet this condition but still are below 1%, which admittedly ensures that REE are indeed rare. In addition, we provide in the Material and Methods section on Modelling and Statistical Analysis, a characterization of the convexity properties of all options, in terms of Jensen gaps that we relate to statistical moments in closed-form.

Panel (d) in Figure 1 is the central graphic tool for presenting the main behavioral data from our experiments. Because, as emphasized so far, convexity of available options is a key dimension in our design about decision-making under REE, our first task is to track how often rats have chosen convex rather than concave options in gains/losses, over the course of the 41 sessions that each of our 20 rats has run. Second, we also aim at assessing whether rats have possibly picked convex options more or less often in the loss domain than in the gain domain, which would depict an asymmetric sensitivity to REE. Both dimensions easily combine in panel (d). On the vertical y-axis we report the sum of the frequencies of choosing both convex gains (that is, Anti-fragile or Vulnerable) and convex losses (that is, Anti-fragile or Robust), while on the horizontal x-axis is depicted the difference, defined as the frequency of picking convex options in the gain domain minus the frequency of choosing convex options in the loss domain. The resulting rotated square in panel (d) of Figure 1 is then a convenient way to report each rat’s choices over one or several sessions, along both dimensions and can be used to identify different types of behavior. More details about how to formally construct such a rotated square are given in Material and Methods, subsection Modelling and Statistical Analysis.

To see why panel (d) is useful, let us first focus on its 4 edges, which represent extreme cases in the sense that they correspond to specialized choices: if an hypothetical rat’s behavior is summarized by a point located at exactly the Anti-fragile edge, it means that this rat has exclusively chosen the Anti-fragile option at each nose poke during all 41 sessions. As a consequence, the vertical y coordinate reaches its maximum value (2 if frequencies are added up) while the horizontal x coordinate is zero since both convex gains and convex losses are - symmetrically - chosen all the time. Following the same logic, the Fragile edge has coordinate {0, 0} since it involves choosing concave gains and losses all the time, i.e. never choosing convex options. The Robust option, in contrast, is depicted by {−1, 1} since it implies picking exclusively convex losses but concave gains, in an asymmetrical fashion. The reverse asymmetry characterizes the Vulnerable edge at {1, 1}, with convex gains and concave losses chosen at all times. Now when a rat’s behavior happens to be represented by a point that is not located at any of the 4 edges, such a location indicates that that rat diversifies (that is, mixes) across options. For example, the green point in panel (d) is close to the Anti-fragile edge and located on the vertical line linking it to the Fragile edge, which might happen if a rat would pick the former with, say, frequency about 95% of the time and the latter about 5% of the time. The red point would be attained by reversing those proportions. Of course many other combinations of the 4 options is possible. For example, the blue point depicts an hypothetical rat that would diversify across the Anti-fragile and Robust options almost equally, at the expense of the two remaining options that have negligible frequencies. Th next section will show that diversifying primarily across two options is indeed the main pattern featured in our behavioral data.

A feature of our experimental design and its implications need to be emphasized in more details at this point. First, the set of NE has total probability of about 2(p1+p2) ≈ 0.9 since ε is a small number. RE are rare but not extreme is the sense that their total probability 2p3 ≈ 0.1 is moderately small while their consequences are moderately large. REE, in contrast, are more extreme outcomes, since they imply waiting 240 seconds (the Black Swan) and gaining 80 pellets (the Jackpot), with a much smaller likelihood. In practical terms, the frequency of each REE is ex-ante around 1% for all rats. Despite their small probability, REE have some importance in the following sense. As shown in Material and Methods, subsection Modelling and Statistical Analysis, the values for both frequencies and outcomes ensure that concave options dominate, in the sense of first-order stochastic dominance, convex options over NE, as well as in the sense of second-order stochastic dominance over the union of NE and RE. In other words, if the extreme gains and losses do not materialize, concave gains (losses) dominate convex gains (losses). However, this does not necessarily hold when REE are added: provided that ε ≥ 0.302% (which we assume hold ex-ante), convex gains (losses) dominate concave gains (losses) in the sense of second-order stochastic dominance. This implies, in particular, that over the full domain that is the union of NE, RE and REE, expected gains for convex options are larger than expected gains for concave options. Likewise, expected time-out punishments for convex options are smaller than expected gains for concave options over the full range including NE, RE and REE.

The reason that we impose such a dominance reversal is as follows. In theory, a rat choosing concave options at all times can be thought of as having absolutely no sensitivity to REE: it acts as if those events never occur and always go for first or second-order stochastic dominance (over NE and RE) as Decision-Making criteria. And those criteria unequivocally point to the Fragile option as the best choice. This gives our first measure along the y-axis of panel (d), Total Sensitivity to Rare and Extreme Events, that again simply sums up the proportion of convex exposures that are chosen for each rat over the 40 sessions. A rat picking the Fragile option at all times has zero Total Sensitivity, since it behaves as if the Black Swan or the Jackpots never occur. At the other extreme, a rat picking exclusively the Anti-fragile option signals maximal Total Sensitivity, since it always stays exposed to the Jackpot while always avoiding the Black Swan. Moderate Total Sensitivity arises, then, if a rat diversifies its choices, for instance primarily across either Anti-fragile and Fragile, or Robust and Vulnerable. Note that such pairs obviously achieve identical outcomes in terms of pellets and time-out punishments when chosen with equal frequencies.

Because gains and losses are deliberately integrated in our design, we also need to distinguish whether rats tend to choose convex exposures symmetrically over the gain and loss domains. This is the purpose of the measure along the x-axis in panel (d). Suppose that we focus for now on the upper half of the square, which implies that Total Sensitivity to REE is larger than one. We then say that a particular rat exhibits Black Swan Avoidance when it picks convex exposures in the loss domain more often than in the gain domain. Symmetrically, Jackpot Seeking is the label we use to depict a situation when convex choices are more often made in the gain domain. Note that these statements, rather qualitative at this stage, will be quantified below. Given its Total Sensitivity larger than one, a rat’s behavior exhibits Black Swan Avoidance (Jackpot Seeking) whenever it is represented by a point that is located on the upper right (left) quarter of the square in panel (d). Symmetrically, for Total Sensitivity smaller than unity, one concludes to Jackpot Avoidance (Black Swan Seeking) whenever the rat is represented by a point that is located on the lower right (left) quarter of the square In other words, the numbers read on the y-axis measure One-Sided Sensitivity to Rare and Extreme Events: when negative (positive) it indicates REE Avoidance (REE Seeking). Fully tooled-up with both measures, we report in the following sections our main results.

Main phenotype features quasi-complete Black Swan Avoidance but partial Jackpot Seeking

In this section, we present the behavioral data coming from the 41 sessions that we have been running for each of the 20 rats - see the Material and Methods section on Experimental Method Details for more details. We organize the data in the following way, so as to underline two features of the choices made by the sample of rats. In each panel of Figure 2 is depicted the (rotated) square that we have already presented in panel (d) of Figure 1, the edges of which represent the four exposures if chosen exclusively at all times. In the square, each point represents a rat. A quick glance at Figure 2 already shows that virtually no such point is located exactly at any of the 4 edges. Quite to the contrary, since it turns out that all rats diversify their choices across a set of options, which is reminiscent of observed behaviors such as, for example, bet-hedging in animals and financial portfolio strategies used by humans (see [5], [36], [39], [44]).

Total Sensitivity to REE (y-axis) against One-sided Sensitivity to REE (x-axis; i.e. REE seeking towards the right and avoiding towards the left) - see Material and Methods section on Modelling and Statistical Analysis - panel (a) black-colored points depict 41 sessions run by 20 rats all together; panel (b) in each square representing a rat, colored points depict 41 sessions for each of the 20 rats; the color code indicates the 2 main categories of profiles observed (blue: the most sensitive to REE and black swan avoiders; green: sensitive to REE but neither biased towards seeking nor avoidance (close to midline on the graph); 2 outliers were identified and color-coded in red and pink; panel (c) color-coded points depict averages over the 41 sessions for each of the 20 rats such points. Most points lie on the right half of the square, that is, on the REE Avoidance side. In addition, within the right half most points are located in the upper quarter, that is, above the Robust-Vulnerable vertex, suggesting moderate to high Total Sensitivity as well as Black Swan Avoidance. In the Material and Methods section on Experimental Method Details, we discuss the contribution of the training sessions to such patterns. These two patterns are better seen if we focus on each individual rat.

How rats diversify their choices can be represented in three steps, focusing first on all choices made by all 20 rats over the 41 sessions, depicted by black points in panel (a) of Figure 2. More precisely, each black dot represents a rat’s session, with a total of about 800 We now look separately at individual behavior for each of the 20 rats across the 41 sessions, in panel (b) of Figure 2. Each rat is represented by a square and, within each square, a point represents one of the 41 sessions for that particular rat. For convenience, the vertices of all squares are labelled by letters: “A” obviously stands for the Anti-fragile exposure, while “R”, “V” and “F” stand respectively for the Robust, Vulnerable and Fragile options. All rats exhibit moderate to high Total Sensitivity, with the exception of rat 17 depicted in red in panel (b) of Figure 2. Among this vastly dominant group, one can identify two sub-groups, color-coded in blue and green, and a second outlier. Starting with the latter, rat 19 (in purple) is the sole rat that exhibits both the highest Total Sensitivity and REE (Jackpot) Seeking, by mixing mostly Anti-fragile (quite often) and Vulnerable (less often) options.

Except for rats 17 and 19, the remaining 18 rats can be gathered in two sub-groups. The largest one gathers the rats colored in blue: rats 1 to 6, 10, 12 to 14, 16, 18 and 20 diversify their choices primarily across the Robust and Anti-Fragile options, with various frequencies. This means that the dominant phenotype, observed over 13 rats depicted in blue, exhibits both moderate to high Total Sensitivity together with Black Swan Avoidance: rats within this group try more often to avoid the Black Swan than to be exposed to the Jackpot. A smaller sub-group composed of the rats depicted in green, however, diversify their choices rather across the Fragile and Anti-Fragile options: those 5 rats depicted in green exhibit also Total Sensitivity but feature neither Black Swan Avoidance nor Jackpot Seeking: those rats have zero One-Sided Sensitivity or, in other terms, make choices that feature symmetry towards in terms of neither avoiding nor seeking REE.

Third, the two groups of blue and green rats are conveniently represented in panel (c) of Figure 2. Each point is associated to a particular rat and it represents the average of its choices over the 4A sessions, as measured through our two metrics. As for Total Sensitivity, one notices that all rats but 4 have an average sensitivity larger than half its maximal value: 3 rats having a Total Sensitivity slightly below 1 and a single rat with a value around 0.5. The latter (rat 17 in red) is the closest to a Fragile behavior, with a significant deviation from it, though. Overall, therefore, most rats (that is, 19 out of 20) are rather sensitive to the presence of REE. In addition, panel (c) confirms the striking feature that most rats exhibit Black Swan Avoidance (i.e. negative One-Sided Sensitivity). Rats 17 and 19 (in red and purple respectively) are outliers for opposite reasons. The latter is the unique rat with low Total Sensitivity while the former is the rat with the largest Total Sensitivity and with pronounced Jackpot Seeking behavior. Among the remaining group composed of 18 rats, all with moderate to high Total Sensitivity, the average behaviors in panel (c) allow one to identify the two sub-groups that we have previously stressed: the 5 rats depicted in green have zero One-Sided Sensitivity (neither Black Swan Avoidance nor Jackpot Seeking) while the 13 blue rats exhibit pronounced Black Swan Avoidance. Exact values can be found in the Supplementary Material section on Behavioral measures for each rat.

An interesting characteristic appears from direct inspection of panel (c) in Figure 2: the sample of 16 rats with the largest Total Sensitivity features a positive relationship between Total and One-Sided Sensitivities. In other words, among those rats that have both moderate to high Total Sensitivity and Black Swan Avoidance, rats that are more sensitive overall to convexity tend to also exhibit lower Black Swan Avoidance (that is, less negative One-Sided Sensitivity). It is as if large Total Sensitivity comes with a mixture of exposures that tend to favor convexity in the gain domain relative to the loss domain. In a sense, rat 19 in purple can be seen as a limiting example of that behavior: it combines the Anti-fragile and Vulnerable options so as to approach near-maximal Total Sensitivity and Jackpot Seeking as opposed to Black Swan Avoidance.

This property is confirmed if we compute the correlation between average Total Sensitivity and average One-Sided Sensitivity. For example, for the whole sample of 20 rats, this correlation is close to (and not statistically different from) zero, while it has a value of about 0.57 over the 41 sessions within the group of 16 rats with Total Sensitivity larger than one. In other words, if we drop the 4 rats with the smallest Total Sensitivity from the sample (those rats represented by points below the Robust-Vulnerable vertex in panel (c) in Figure 2), we are left with 16 rats for which larger Total Sensitivity is associated with lower One-Sided Sensitivity/less Black Swan Avoidance, that is, more symmetry and even Jackpot Seeking for rat 19. Moreover, if we divide the 41 sessions of the 16-rat group over time, the correlation between Total and One-Sided Sensitivity increases from about 0.17 on average in the first 10 sessions, to 0.56 in the next 10 sessions, to 0.71 in the next 10 sessions, and finally to 0.64 in the final set of 10 sessions. Therefore, the pattern we alluded to appears early in the sessions and is consistent over time.

In Figure 3, we additionally report behavioral measures of Black Swan Avoidance and of Jackpot Seeking, defined as the shares of nosepokes that respectively lead to exposure to Jackpots and non-exposure to - i.e. avoidance of - Black Swans (how those behavioral measures relate to Total and One-Sided Sensitivities is described in details in the Material and Methods section on Modelling and Statistical Analysis). What the median values of Black Swan Avoidance and of Jackpot Seeking that appear in Figure 3 reveal is that the blue and green phenotypes differ sharply in one respect: half of the rats depicted in blue achieve nearly complete Black Swan Avoidance since more than about 90% of their nosepokes lead to avoid being exposed to Black Swans, compared to more than about 40% for half of the green rats as well as for the red rat and pink rat. In contrast, blue and green rats do not differ in terms of median Jackpot Seeking. This is confirmed by paired Wilcoxon tests: Black Swan Avoidance and Jackpot Seeking are significantly different for blue rats (p = .0002) but not for green rats (p = .6250). The two rats classified as outliers (red and pink) differ from both blue and green rats in terms of Jackpot Seeking. The red rat seeks quasi-complete exposure to Jackpots whereas the pink rat seeks quasi-complete avoidance of Jackpots. Exact values can be found in the Supplementary Material section on Behavioral measure for each rat.

Median behavioral measures of Black Swan Avoidance and Jackpot Seeking by phenotype, defined as median shares of nosepokes exposed to REE (in percentage) for blue and green phenotypes; phenotypes are indicated in column with a corresponding colored symbol at the center - left panel depicts the percentage of nosepokes exposed to Black Swans in black and not exposed to Black Swans in light gray - Black Swan Avoidance increases with the share of non exposure to Black Swans; right panel depicts the percentage of nosepokes exposed to Jackpots in black and not exposed to Jackpots in light gray - Jackpot Seeking increases with the share of exposure to Jackpots; blue and green phenotype comparisons are carried out by Wilcoxon Rank Sum tests (⋆ means p < 0.05 and ⋆⋆ means p < 0.01)

Black Swans do reinforce sensitivities to REE while Jackpots do not

Having documented that most rats (19 out of 20) exhibit medium to high Total Sensitivity and medium to high Black Swan Avoidance, we now address the issue of how rats respond to the actual occurrence of REE within a session. More precisely, we aim at comparing the average behavior of each rat before and after the occurrence of Black Swans and Jackpots. In Figure 4, we report how REE affect Total and One-Sided Sensitivities, within a window formed by the average over 10 nose pokes before and the average over 10 nose pokes after. For both panels (a) and (b), each point represents a particular rat and the colors group all rats as in panels (b)-(c) of Figure 2. In panel (a) of Figure 4 we report the impact of Jackpots while in panel (b) is depicted the effect of Black Swans, for Total Sensitivity on the left and One-Sided on the right. Dotted 45-degree lines represent the no-effect benchmark.

Responses of Total Sensitivity to REE and of One-sided Sensitivity to REE, averaged over the 41 final sessions, following Jackpots in panel (a) and Black Swans in panel (b), for each of the 20 rats. Each dot indicates one rat and dots are color coded to indicate the profile of the animal, as in Figure 2. Black dotted lines materialize no change in sensitivities after exposure to the REE compared to before. Black solid lines represent spline estimates - see Material and Methods section on Modelling and Statistical Analysis

Visual inspection of both panels suggests that while Jackpots tend not to affect both sensitivities much, Black Swans do and tend to increase Total Sensitivity and to decrease One-sided Sensitivity: most dots do tend to be located above (below) the 45-degree line in the lower left (right) panel. In other words, rare and extreme losses/Black Swans tend to reinforce Total Sensitivity, contrary to extreme gains/Jackpots. This reinforcement effect due to experiencing negative REE is consistent with our main result that most rats are sensitive and chose to avoid extreme losses, that is, exhibit medium to high Total Sensitivity and Black Swan Avoidance. In addition, panel (b) also reveals that rare and extreme losses tend to make Black Swan Avoidance stronger: most dots tend to be located below the 45-degree line, while in the lower right panel. Again, this is consistent with the observation that extreme losses are somewhat given more weights than extreme gains by most rats, which implies that Black Swan Avoidance is reinforced when a extreme loss is experienced. This is confirmed by statistical tests, including one using bootstrap, as shown in the Material and Methods section on Modelling and Statistical Analysis (see in particular the discussion of Table 3). In sum, a within-session analysis of the effects of experiencing a REE shows that the occurrence of Black Swans tends, in the short-run, to significantly reinforce both Total Sensitivity and Black Swan Avoidance, while the occurrence of Jackpots tend to leave Total and One-Sided Sensitivity unaltered.

Q-Learning models require specific REE decision weights to mimic rats’ behavior

We have introduced sensitivity to Black Swans and Jackpots in a new class of augmented Reinforcement Learning models and we have estimated their parameters using observed choices and outcomes for each rat. The selected model ended up with a distinctive feature: it separates normal from rare and extreme outcomes through different weights in the Decision-Making process. Adding such specific sensitivity results in a good fit of the selected model and simulated behaviors that are close - to behavioral observations, whereas a standard Q-learning model without sensitivity to REE is rejected for almost all rats.

To each of the four available options, indexed below by “o” (from 1 to 4, referring to Antifragile, Fragile, Robust, Vulnerable) in equations (1), we attach at each moment in trial/time t a gain sub-value Qg(o) and a loss sub-value Ql(o) that are updated as follows: where the convention that updated values are indexed by t + 1 means right after observing the gain or loss corresponding to the choice made at t. Parameters αg and αl are usually labeled learning rates, indicating the speed at which reward prediction error gets updated into the latest values. Limiting cases occur when setting the learning rate either to zero - its smallest possible value - which means no subvalue updating, or to one - its largest possible value - which means that the subvalue tracks the obtained reward itself. While equations (1) applies for the updating of the chosen option, a similar updating rule applies for the three other options that are not chosen at that moment, by assuming that this happens as if the reward was zero, with different parameters. That is, for options that are not chosen, the updating rules for subvalues are simply and , where the forgetting rates and are assumed to be bounded between zero and one.

The novel feature of our augmented Q-Learning model, besides integrating gains and losses, is that specific weights are attached to options that may produce REE. More specifically, we model the value of each option o as:

In equation (2), the value of each option Vo is the sum of two terms. The first is the decision weight attached that sums up the gain and loss subvalues, each weighted by parameters λg and λl that may reflect a differential effect. Importantly, in the class of nested models that we estimated, REE may or may not be incorporated in the gain and loss subvalues, that should be thought of as averages - though not arithmetic but of an exponential moving type in view of equations (1) - see page 32 in [46]. The second and key term in equation (2) specifically captures the decision weight attached to REE, and it can itself be decomposed into the decision weights on Jackpots and on Black Swans. More precisely, the indicator function 1JP (o) equals one if the option exposes to Jackpots (namely when the chosen option is either A or V) and zero otherwise. Similarly, 1BS(o) equals one if the option exposes to Black Swans (namely options F and V) and zero otherwise. Therefore, γg and γl reflect the (possibly different) weights attached to Jackpots and Black Swans in the decision to choose one particular option rather than any other available.

Finally, the associated so-called action probabilities are given, for each option, by the Softmax function (a.k.a. the multinomial Logit model) , each a number between zero and one. In the estimation procedure detailed below, the parameters of interest are γg and γl in the sense that the benchmark model sets them to zero. Our candidate models are then compared against such a benchmark in two ways: one sets of models incorporates REE into sub-values while the second does not, and γg and γl are estimated for both. We hope it has become clear by now that the interest of the latter set of models is of interest because it indicates, if selected, that REE are integrated into the Decision-Making process in a separate way and not through averages, that is, not through the sub-values. In sum, the class of augmented Q-Learning models with gains and losses has at most 8 parameters: two learning rates, two forgetting rates, two decision weights for sub-values and two decision weights for REE.

We have used information criteria to determine the selected model for each rat over all its choices, as summarized in Table 1 by phenotype. Selected models indicate that 17 rats out of 20 integrate REE, at least using specific decision weights, in their Decision Making process, while only the remaining three do not. Most animals (8 blue rats, 4 green rats, as well as pink and red rats) integrate REE only though their specific decision weights (with or without specific forgetting parameters). Three blue rats integrate REE both with specific decision weights and in Q-sub-values. In contrast, the learning and forgetting parameters do not help to discriminate models.

Figure 5 reports the predicted (bottom left) and simulated sensitivities (bottom right), which can be compared to actual sensitivities (top, which replicates the right panel in Figure 2). The predicted sensitivities, generated using the estimated parameters of the selected model given the actual rewards obtained during the 41 sessions, are accurate descriptions of the observed sensitivities computed from the actual choices. Simulated sensitivities are obtained by running the selected model over artificially generated 41 sessions, using its median parameter values for blue and green rats (and individual parameter values for the pink and red rats). These simulations show first that the selected Q-Learning model generate behavioral data that are stable for the blue rats as well as for the red and pink ones. Second, simulated green rats tend to separate into two stable groups, still along the vertical line.

Total Sensitivity to REE (y-axis) against One-sided Sensitivity to REE (x-axis) averages for each rat over the 41 sessions; top panel replicates the right panel in Figure 2: observed sensitivities; bottom left panel: predicted sensitivities derived from selected models; bottom right panel: simulated sensitivities computed from runs of selected models over artificial 41 sessions

Finally, estimated parameter values appear in Figure 6, color-coded in the same way as in previous figures. The details about nested models and how the selected one is obtained are presented in the section of the Materials and Methods on Augmented Q-Learning Model Estimation and Simulation. The main result in Figure 6 relates to a major difference between the behaviors of blue and green rats, as seen through the lens of our augmented Q-Learning model: blue rats have a significant specific sensitivity to Black Swans that is translated into a very negative decision weights attached to options that expose to them (p = 0.0102). In other words, blue rats have a negative γl, which means that they tend to avoid the options with Blacks Swans (F and V) most of the time. Along that dimension, they differ significantly from green rats, which mostly mix options F and A.

Parameters of the selected Q-Learning model, estimated for each rat over the 41 sessions - left panel: decision weights for averages of gains and losses and for REE; right panel: learning and forgetting rates for gains and losses; while individual parameter for pink and red rats are represented as dots, parameters for blue and green rats are represented by box plots with 10th and 90th percentiles, interquartile ranges and median; between phenotype comparisons carried out by Wilcoxon Rank Sum tests (⋆ means p < 0.05) - see Material and Methods subsection Augmented Q-Learning Model Estimation and Simulation for details

Discussion

Quasi-complete Black Swan Avoidance vs partial Jackpot Seeking as evidence that extreme gains and extreme losses are treated differently

To sum up the behavioral results: we have shown, first, that all rats in our sample exhibit some Total Sensitivity - in fact, medium to large Total Sensitivity for most rats - to REE. And, second, that the main phenotype features nearly complete Black Swan Avoidance coupled with partial Jackpot Seeking. We interpret these results as intimately related to a fundamental asymmetry between negative and positive REE. In our design, rats have to endure larger frequent delays to avoid Black Swans and to forgo larger frequent gains for a Jackpot to occur. However, rats can avoid with certainty the Black Swans when doing so, making convex menus over losses relatively more attractive, but cannot guarantee that a Jackpot will occur, making convex options over gains relatively less attractive. The actual occurrence of the latter will depend on the particular draw obtained in each of the repeated sessions. As an illustrative example, suppose one uses a Poisson approximation of the probability that the REE materializes given that 150 choices are made within a session, which falls within the range of the number of nose-pokes that rats have performed. If one further supposes that the ex-ante probability of REE is 1%, then the probability to observe exactly one REE in that session is about 78%, which leaves some room for the REE to not occur. Our design thus captures an asymmetry that arguably distinguishes positive from negative REE in general: while it may be possible to completely avoid the latter, the former is never expected to occur for sure even if the appropriate choices are repeated. In fact, the rarer the Jackpot, the larger the number of trials needed to experience at least one, requiring marked perseverance in Jackpot Seeking.

In the specific context of our experimental task, this asymmetry implies that although complete Black Swan Avoidance means avoiding Black Swans with certainty, complete Jackpot Seeking does not ensure that Jackpots will necessarily occur. In sharp contrast, the values corresponding to the frequent domain - i.e. the union of NE and RE - will be observed often given their ex-ante probabilities larger than 10%. In other words, while inference about how often REE - and Jackpots for that matter - occur is not reliable, accurate inference about how the frequent gains materialize seems much more within reach given the many choices made in the conditioning and final sessions.

This feature of the experimental design has important consequences for the choices that rats make. To the extent that they explore different options, rats are subjected to a flood of stimuli that point at concave options as the optimal choice: the Fragile option, in particular, generate larger gains but smaller losses in the frequent domain and in fact it stochastically dominates all the others if REE do not materialize. To put it differently, for each session, the Fragile option is dominated by the Anti-fragile one if both the Black Swan and the Jackpot occur. However, because complete Black Swan Avoidance practically insures against the occurrence of Black Swans, it seems like an attractive combination even if this implies violating Stochastic Dominance in the frequent domain: Stochastic Dominance over the full domain is attained with certainty. In contrast, complete Jackpot Seeking is less attractive since it does not ensure that a Jackpot will occur: the cost of getting smaller gains most of the time means that partial Jackpot Seeking makes sense as a good compromise since it implies that Stochastic Dominance in the frequent domain is less often violated, in a context where Stochastic Dominance over the full domain cannot be attained with certainty.

In sum, bearing the cost of getting smaller gains frequently is acceptable only if the compensating benefit in the form of a Jackpot materializes when many choices are made. Similarly, bearing the cost of getting larger losses frequently is acceptable only if the compensating benefit in the form of an avoided Black Swan is ensured. While the latter condition is certainly met if Black Swan Avoidance is complete, the former condition is not even if Jackpot Seeking is complete. Results suggest that rats depicted in blue have learned this property, since they combine mostly the Robust and Anti-fragile options to obtain quasi-complete Black Swan Avoidance together with partial Jackpot Seeking. In contrast, the rats depicted in green seem not to have learned this property, as they mix Anti-fragile and Fragile options, thus remaining exposed to Black Swans.

In theory, since frequent outcomes occur often when choices are repeated many times, Stochastic Dominance in the frequent domain should be detected and learned rather quickly by animals. However, the experimental literature shows that animals need a large number of repeated sessions to learn Stochastic Dominance (e.g. [55], [3]). We think our results suggest a potential explanation : animals may need a significant amount of sampling to infer that REE are excluded. In addition, one may wonder at this stage whether our results can be expressed in terms of loss - and possibly ambiguity - aversion. Although most other studies have not focused on REE, to the best of our knowledge, neuroeconomic tasks using positive and negative outcomes have generally shown that loss aversion occurs (see e.g. [17], [13], [57]). This bias towards negative/losses in the valuation of outcomes has been previously shown and is in line with Prospect Theory ([21]), although models following this approach fit better rat’s behavior in tasks only involving positive rewards and in which loss is not a punishment, but rather a lack of positive reward ([10],[3]). It could also include some emotional/anxiety-related component ([56]), possibly involving the basolateral amygdala as shown in rats ([50]). Ambiguity aversion in animals, on the other hand, seems to have been less explored (see [40] for non-human primates). Our results tentatively suggest that rats might conform to ambiguity reduction regarding REE, when combining Black Swan Avoidance that is quasi-complete and Jackpot Seeking that is only partial: while the Jackpot’s sampling frequency of occurrence is in each session bounded below by zero and above by ε at best, the Black Swan’s sampling frequency approaches zero when Black Swan Avoidance is nearly complete.

The quasi-complete Black Swan Avoidance together with partial Jackpot seeking, that we report for the main phenotype represented by rats depicted in blue, provides further evidence that extreme gains and losses receive differential treatment in the Decision-Making process. However, we show that the classical Q-Learning models which account for a potentially different treatment of gains and losses do not fit our behavioural data. Interestingly enough, observed behaviors are in contrast neatly captured in the selected Q-Learning models that are augmented with specific sensitivities to REE, and precisely through a marked avoidance to Black Swans which sets the main phenotype represented by rats depicted in blue apart from rats depicted in green.

The augmented Q-Learning models suggest a specific cerebral pathway for REE

Both the good fit and the accurate simulations of the augmented Q-Learning model for the main behavioral phenotype suggest a novel neurobiological conjecture. Viewed from the lens of the model, 11 rats (out of the 13 that form the blue group, as detailed in the Material and Methods section) exclude REE for the Q sub-values while expressing a specific sensitivity to Black Swans through a negative parameter γl. This means that when evaluating available options, “blue” rats do not incorporate Black Swans into the average of losses but rather attribute them a specific decision weight, with the effect that options exposing to Black Swans become less attractive. For humans, extreme stimuli have been shown to be more psychologically salient in perception and memory than moderate stimuli in [25]. Our experimental and modelling results suggest that something similar could happen for rats. This insight from modelling opens up the possibility that rare and extreme outcomes, especially rare and extreme losses, could be encoded differently in the brain, compared to frequent and smaller rewards that are classically considered to be encoded by the DA neurons ([9], [42], [57], [51]), the striatum ([7], [11]), and cortical areas such as cingulate and orbitofrontal cortex ([41], [14])

Further studies are necessary to identify how REE could be encoded and where. If we consider that taking REE into account would require to integrating long-term outcomes and not only short term considerations as studied by [52], we might hypothesize that the regions of interest could be the anterior insula and amygdala. The latter being well known for its involvement in emotional processes, it is wise to consider it could be involved in REE-related events, which are highly salient and emotionally charged, and affect Decision-Making with a differential treatment of negative outcomes ([7], [50]). From a statistical viewpoint, it makes sense not to add up “outliers” into the first moment - the mean - of the observed distribution, because doing so could result in an unstable sample average. In a related way, good Decision-Making might depend on giving special treatment to REE over frequent rewards - as well as differential treatment of Black Swans vs Jackpots - and if such practice is evolutionary advantageous in uncertain environments (see e.g. [8]), this opens up the possibility that this would be reflected in the brain, possibly in different species. To the best of our knowledge, such an hypothesis that builts upon experimental and modelling results such as the ones we report here has not been put to the test in neurosciences, as it should in view of the arguments developed above for humans and other animals. This aspect could in principle be addressed within a computational validity framework, as advocated in [38].

Relevance for humans

Whether humans would find Black Swan Avoidance and Jackpot Seeking attractive and how they would combine them in a related experimental setup is a question that comes naturally to mind in view of the results reported in this paper. Our novel experimental design can in fact be adapted to human participants, so as to test in a similar fashion whether they adopt behavior that tend to avoid harmful - and seek beneficial - REE. This makes the present results obtained with rats also relevant for pressing questions that link the past and future of humanity. Usually gathered under the concept of the Anthropocene, there is mounting evidence that climate and other environmental changes mainly driven by human activities originate extreme events (see e.g. [12]), raising the possibility of rare ones that could lead to partial or full extinction for many species. Even though sizable uncertainty remains about the magnitudes involved, the associated empirical evidence puts the human species in a specific situation that raises difficult but pressing questions. Why has the human species not succeeded in the past in avoiding courses of action that could potentially lead to such harmful REE, and to what extent will humans be able to cope with them in the near future, should they happen? In other words, are humans engaged in a path that diverges from Black Swan Avoidance instead of approaching it?

While intuitive, the notion that some adaptive and survival value derives from animal behaviors that help avoid destructive REE, or at the very least help cope with them should they happen, still lacks compelling evidence (see, however, [8]). If correct, such a conjecture would raise a paradox, given the specific contribution of humans in creating environments that are more prone to harmful REE: do humans possess a special trait that somehow prevented them to insure against such REE, despite the evolutionary advantage possibly attached to avoiding them? Put differently, to the extent that destructive REE are less exogenous to humans, due to their unique set of technological capabilities, than they are to other species, could there exist an epigenetic rule that is specifically human and that is conducive to behaviors that are unable to rule out extremely harmful events? If that is the case, could such an epigenetic rule manifest itself in such a way that the capacity of humans to cope with such REE, if they have become unavoidable, is hampered? For example, is it the case that detecting somewhat predictable cues for REE, such as convexity in our setting, is out of reach (see [2])?

In this study, we report results obtained from behavioral experiments conducted on rats that do not go, fortunately, in the direction of a positive response to the above set of questions. More precisely, the heterogeneity in individual behavior that we observe in rats, when confronted with alternative options in an uncertain context, suggests that some animals may individually lack the ability to insure themselves against a rare but consequential loss, even if they have the possibility to do so. In that sense, the human species, if afflicted by the same incapacity, is not alone in the animal kingdom. Of course, the bright side of the coin is that most rats consistently chose options that avoid almost completely exposure to a harmful REE and even provide partial exposure to beneficial REE in the form of a large gain. As a matter of fact, this group is the most numerous in our sample of rats that have run the experiments. A speculative hypothesis then comes to mind: that some human individuals, if not all, may also settle on behaviors that avoid harmful - and possibly seek beneficial - REE. More generally, if the sensitivity and reaction to REE differ among individuals in some species, this raises the important question of how such heterogeneity affects - and interacts with - population dynamics and social institutions, most importantly collective Decision-Making.

Material and Methods

Experimental Model and Subject Details

Animal Subjects

Adult Lister Hooded males (n = 20, 200 g at arrival, Charles River) were housed in groups of two in Plexiglas cages and maintained on an inverted 12 h light/dark cycle (light onset at 7 pm) with water available ad libitum, in a temperature - and humidity - controlled environment. Food was slightly restricted ( 80% of daily intake). Animal care and use conformed to the French regulation (Decree 2010-118) and were approved by local ethic committee and the French Ministry of Agriculture under #03129.01.

Experimental Method Details

Apparatus

All behavioral experiments took place during the animals’ dark phase in standard five-hole operant boxes (MedAssociates) located in ventilated sound-attenuating cubicles. One side of each box was equipped with a central house light, a tone generator and a food magazine, outfitted an infrared beam for detecting nose poke inputs. Sucrose pellets (20 mg; Bio-Concept Scientific) were delivered from an external food pellet dispenser. An array of five response holes was located on the opposite curved wall, each equipped with stimulus lights and infrared beams for detecting input (nose poke). The center hole was continuously closed throughout the experiments (Fig. 1a). Data were acquired on a PC running MedPC-IV.

Design of Behavioral Sequences

Four menus were elaborated by mixing convex and concave exposures for both the gains (sugar pellets) and the losses (time-out punishment) described on Figure 1 (panel (b)). For the gains, animals could obtain 1 , 3 (NE domain, blue), 12 (RE domain, yellow) or 80 (REE domain; pink) sugar pellets in the convex exposure or 2, 4, 5 or 5 pellets in the concave exposure. For the losses , convex exposure may impose 6, 12 , 15 or 15 sec of time-out punishment while it was 3, 9 (NE domain; blue), 36 (RE domain, yellow) or 240 sec (REE; pink) for the concave exposure.

The four behavioral options depicted in Figure 1, panel (c), are therefore combinations of the above convex and concave exposures: the “Anti-fragile” exposure at the top middle is convex in both gains and losses, while the “Robust” option (left) is only convex for the losses. On the other hand, the “Vulnerable” exposure (right) is convex only for the gains, while the “Fragile” option (bottom middle) is concave for both gains and losses. This implies that the extreme - but rare - gain of 80 pellets (i.e. Jackpot) may be delivered only when either the Anti-fragile or the Vulnerable options are picked, while the extreme and rare time-out punishment of 240 sec (i.e. Black Swan) may be only experienced if choosing either the Fragile or the Vulnerable options.

The first three events of each behavioral options belong to the frequent domain as their frequency of occurrence, respectively 0.5, 0.4 or 0.1, is significantly larger than zero. During behavioral testing, animals equivalently experienced the gain and loss domains. On the other hand, extreme outcomes, i.e. Jackpot or Black Swan, have a much smaller likelihood of occurring since they may appear only at particular point during the behavioral sequences (see below). This means that they could happen if a rat has chosen an exposure that is either convex in the gain domain or concave in the loss domain at a given time. This implies that the frequency of rare and extreme events is less than half of one percent for all rats. Despite their low probability, extreme events have some importance because of their value. Our calibration of both frequencies and outcomes ensures that concave exposures dominate convex exposures: if the extreme gain (Jackpot) does not materialize, the expected payoff in sugar pellets is larger for concave exposures than convex exposures (i.e. ex-post first order stochastic dominance, over the frequent domain). Similarly, the expected time-out punishment is lower for concave exposures compared to convex ones, if the Black Swan does not happen. However, ex-post first-order stochastic dominance is reversed in the presence of extreme events, in which case convex exposures become more interesting in terms of payoff than concave ones. The reason we imposed such a dominance reversal was as follows. If rats always choose concave exposures, then they show no sensitivity to rare and extreme events, since they act as if those events never occur and always go for first-order stochastic dominance. This gives us our first measure, Total Sensitivity to REE, that simply sums up the proportion of convex exposures that are chosen for each rat over the 41 sessions. Because we deliberately integrate gains and losses in our design, we need to distinguish whether rats tend to choose convex exposures symmetrically over the gain and loss domains. We say that a particular rat exhibits Black Swan avoidance when it picks convex exposures in the loss domain more often than in the gain domain. Likewise, Jackpot Seeking occurs when convex choices are more frequent in the gain domain.

To avoid potential learning of the event occurrence during behavioral training and testing, ten different sequences of events, with respect of the first-order stochastic dominance as well as the balance between gain and loss domain exposures, were generated and randomly used for behavioral training and testing (Supplemental Figure S3). Furthermore, to increase the rarity and the unpredictable nature of extreme events, the sequences of events used during behavioral training and testing were declined into seven various sequences, in which Jackpot and Black Swan are either unavailable, solely or both available at a given time point of the sequence of events (Supplemental Table S1): when available, extreme events could be obtained at the 10th or the 60th activation within a given sequence, but could not occur at the same time. For example, in a behavioral sequence where the Jackpot should be available at the 10th position, any poke in the Anti-fragile and Vulnerable holes following nine responses made in one of these menus (regardless of the positive or negative outcomes) would trigger the delivery of the Jackpot. Of note, depending on the sequence used, animals could experience both extreme events in a single session.

Behavioral Training and Testing

Training was divided into five distinct phases before the final test: acquisition of the food collecting responses, acquisition of nose poking in the holes, training with four holes, attribution of menus to hole and training on the menus (Figure 7). Each session started with the illumination of the house light.

Time-line of experiments

Acquisition of the food collecting responses: Animals were trained to collect sucrose pellets in the food magazine during three 30-min daily sessions (100 pellets max) under fixed ratio 1 schedule of reinforcement (FR1): one nose-poke into the food magazine triggered the delivery of one sucrose pellet. During this initial phase, nose-poke in any hole had no programmed consequences.

Acquisition of nose-poking in the holes: Here, animals had only access to one hole (the three others being occluded) during the 20-min sessions. One nose-poke in the available hole triggered the illumination of the hole-light and the delivery of one sucrose pellet in the food magazine. Perseverative nose-pokes (those performed before food collection) had no consequences. Following food collection in the food tray, animals were allowed to poke again in the opened hole. During each session, a maximum of 100 pellets were delivered. All animals were trained twice on each hole.

Training with four holes: Following the eight training sessions, animals were allowed to poke in the four different holes during twenty daily sessions of 20-min. Nose poke in any hole triggered the illumination of the associated light and the delivery of one sugar pellet in the food magazine. Both perseverative activations and pokes in other holes had no consequences. After food collection in the magazine, animals were allowed to nose poke again in any hole. The first ten training sessions were limited to 100 sugar pellets. During the following ten sessions, animals were able to collect up to 200 pellets per session.

Attribution of menus to hole: We determined the spatial preference for each rat by establishing the percentage of activation of each hole during the last ten sessions. To favor the emergence of Anti-fragile choices, menus’ attribution was made as follow:

Anti-fragile exposure was associated to the preferred hole

Robust exposure was associated to the 2nd preferred hole

Vulnerable exposure was associated to the 3rd preferred hole

Fragile exposure was associated to the least preferred hole

Training on the exposures: Here, animals were first trained on the gain domain, i.e. no time-out punishment, for each menu/hole. They were subjected to two 20-min sessions, with unlimited number of pellets, during which only one hole was available (eight training sessions in total). For Anti-fragile and Vulnerable options, we used the two sequence-types where the jackpot was available at the 10th and 60th activations (Supplemental Table S1) to ensure that all individuals could experience both an early and delayed jackpot during training. Animals were then allowed to explore all gain options (four opened holes) during 9 20-min sessions, for which different behavioral sequence-types were used. Before training on the loss domain of the different menus, animals were first exposed to a mild and constant 3-sec time-out punishment. Here, animals had access to all options, but half of the activations lead to a 3-sec time-out punishment, notified by a 3-sec tone and the extinction of the house light. During this period, pokes in the different holes or the food magazine had no programmed consequences. After the 3-sec time-out punishment, the house light was turned on and the animals could again pokes in the different holes. Following nine 20-min training sessions, the loss domain of each menus was progressively introduced. As described above, each time-out punishment was notified by a 3-sec tone and the extinction of the house light for the whole duration of the punishment, which termination was signaled by house light illumination. Animals were first exposed to concave exposures, having only access to vulnerable and fragile holes during four 20-min sessions. They were then exposed to convex losses (only Anti-fragile and Robust holes available) for another four 20-min sessions. Thus, at the end of the training, all animals experienced both extreme events at least four times.

Final tests: For the final tests, animals were free to explore all menus throughout the forty 20-min sessions. Animals experienced four times the ten different sequences, randomly distributed across sessions. Population (n = 20) was subdivided into two groups that experienced different, but equivalent, sequence-type distribution (Supplemental Figure S2)).

Figure 7 depicts a simplified time-line of the experimental procedure.

Figure S3 shows the chain of events in the ten sequences used for behavioral training and testing. Numbers in the second column indicate which event would occur and the sign preceding it whether it belongs to the gain (no sign) or loss (minus sign) domains. For example, the third event in the first sequence (noted -2) triggers a time-out punishment of 9 or 12 sec, depending on whether an animal performed his third nose poke in concave or convex menu, respectively. Note that extreme events (which should be noted 4 or -4) do not appear in the sequences, as they would automatically replace the 10th or 60th events of the sequence. See supplementary material for detailed information on sequences of events.

Modelling and Statistical Analysis

Modelling Convex and Concave Exposures

Central to our experimental design is the notion of convex/concave exposure under radical uncertainty, that is, when probabilities and consequences are unknown a priori to subjects. A well known measure of convexity is the Jensen gap that is derived from Jensen’s inequality (see [28] for a graphical exposition), which we now define and relate to statistical moments. In our context, the relevant form of Jensen’s equality states, loosely speaking, that the expectation of a convex function of a random variable is larger than the value of that function when evaluated at the expectation of the random variable. Jensen’s gap is then defined as the difference between the expectation of the function minus the value of the function at the expectation (hence positive by construction). The inequality is reversed for a concave function and the (positive again) Jensen’s gap is then defined as the difference between the value of the function at the expectation minus the expectation of the function. In the gain domain, rats obtain 1, 3, 12 or 80 sugar pellets if the convex exposure is chosen, or 2, 4, 5, 5 pellets if the concave exposure chosen. In the loss domain, convex exposure imposes 6, 12, 15 or 15 seconds of time-out punishment, and 3, 9, 36, 240 seconds for the concave exposure. Because the values in the loss domain are proportional to the values in the gain domain, the former corresponding to 3 times the latter, we focus here on values for pellets. The statistical properties of convex and concave losses follow accordingly. More formally, sequences of gains are ordered by increasing values and are denoted for the convex exposure and for concave gains. We assume identical probabilities for both concave and convex gains, where ε is the ex-ante probability of the rare and extreme event (REE for short). The third value (that is, 12 for convex gains and 5 for concave gains) is labeled a rare event (RE), and the sets of the lowest two values (that is, 1 and 3 for convex gains, 2 and 4 for concave gains) for both exposures are composed of normal events (NE).

Making use of the exponential transform, we define Jensen gaps for the convex and concave exposures as, respectively, which relate to statistical moments as follows. The moment generating function for the convex and concave exposures are given by, respectively: . The series expansion of the exponential function allows us to derive: where and are the N -th statistical raw moments of the convex and concave exposures, respectively. For example, for any integer N ≥ 1. Jensen’s gaps for both exposures are then given by . Using again the series expansion of the exponential of the first moment for each exposure, it follows that both positive Jensen gaps can be written in terms of moments, as follows:

Note that the sums in equation (3) nicely allow for a straightforward decomposition in terms on all raw moments with order larger than two.

More specifically, we set probabilities to identically for both concave and convex gains. Treating the probability for the REE as a varying parameter, it can be shown that Jensen’s gaps for convex and concave exposures are ordered, with the former larger than the latter when ε ≤ 0.02. Over that range, it also holds true that both Jensen’s gaps are monotone increasing function of ε. While this property also holds for the expectations and variances of both exposures, it doesn’t for higher-order moments. For example, the skewness and kurtosis of the convex exposure turn out to be hump-shaped functions of ε, with peaks corresponding to values of ε smaller than one percent. This fact underlines that convexity measured by Jensen’s gap offers a unifying approach to rank exposures, seen as lotteries, whereas statistical moments do not necessarily do so. We conjecture that, more generally, all lotteries satisfying our assumptions (including monotone probability distribution) can be ranked according to their convexity as measured by the Jensen gap.

In Table 2, we report the central moments and the log of Jensen gaps in the top five rows, for both exposures when ε = 1/75 ≈ 1.3%, which corresponds to obtaining one REE out of 75 nose pokes. For that particular parametrization, convex and concave exposures differ mostly in their respective kurtosis and in their Jensen gaps. From row six and below, we report the decomposition of Jensen gaps in terms of the raw moments. For example, in row six is given the ratio of (m2 − (m1)2)/2 - that is, half the variance - to the corresponding Jensen gap for each exposure, in percentage. Strikingly, while moments of order 2 to 8 explain about 90% of the concave exposure’s Jensen gap, they contribute negligibly to that of the convex exposure. In fact, it turns out that moments with order around 80 start contributing, although for a small share each, to the convex exposure’s Jensen gap. In sum, while for the concave exposure a few low-order moments concentrate the contributions to the Jensen gap, the contributions of raw moments to the convex exposure’s Jensen gap spread across a larger number of very high order raw moments.

We derive next the ex-ante properties of the probability distributions associated with concave and convex gains (for definitions of stochastic dominance and congruent utility classes, see for example Fishburn and Vickson [15]; also note that first-order, resp. second-order, stochastic dominance implies second-order, resp. third-order, stochastic dominance):

In the domain restricted to NE, concave gains first-order stochastically dominate convex gains. In addition, concave gains then have a larger expected value than that of convex gains, with equal variance, skewness and kurtosis for concave and convex exposures.

In the domain restricted to NE and RE, concave gains second-order stochastically dominate convex gains. In addition, concave gains then have a larger expected value and smaller variance, skewness and kurtosis than that of convex gains.

In the full domain including NE, RE and REE, convex gains second-order stochastically dominate concave gains if and only if ε ≥ 0.302%. In addition, convex gains then have a larger expected value, variance, skewness and kurtosis than that of concave gains if ε ≥ 0.302%.

The above assumptions on the experimental design are stated in terms of stochastic dominance and moments of the probability distributions. They also have implications in relation to standard approaches to decision-making. First, from an ex-ante perspective with perfect information about the probability distributions of all exposures, value-maximizing subjects whose preferences are represented by any non-decreasing value function would choose concave gains in the domain restricted to NE. They would continue to do so in the domain restricted to NE and RE for any non-decreasing and concave value function. However, in the full domain with REE, value-maximizing subjects endowed with any non-decreasing and concave value function are predicted to choose convex gains.

We denote the value attributed to sequences of gains with probabilities , for n ≤ 4. The above assumptions have the following implications in terms of value maximization:

In the domain restricted to NE, holds.

In the domain restricted to NE and RE, holds.

In the unrestricted domain, holds if and only if ε ≥ 0.302%.

Note that the reversal in second-order stochastic dominance (and value) in the full domain favors convex gains when extreme events that are indeed very rare - with probabilities much smaller that one percent - are added. In contrast, absent REE, adding only a RE is not enough to favor convex gains, even though it allows subjects to possibly detect convexity/acceleration and concavity/deceleration of gains and losses.

The above assumptions hold true under expected utility, that is when with the appropriate conditions on the utility function u (non-decreasing for first-order stochastic dominance, non-decreasing and concave for second-order stochastic dominance; see [15]). To better match experimental data, however, expected utility is increasingly supplemented by some form of probability weighting w(pi). For instance, [10] use a two-parameter functional form due to [35] and find that 30 rats out of 36 behave as if their probability weighting w(pi) is concave. In that case, the above ex-ante properties hold mutatis mutandis, for example, under expected utility with probability weighting, defined as . Alternatively, in the setting of rank-dependent expected utility, we can make use of results 3 and 4 in [24] to show that the above ex-ante properties still hold provided that the transformation of the cumulative distribution function is any increasing-concave function. Note that since in our experimental design REEs are both extremes (in the sense of being the largest values) and rare (that is, they have very low probabilities), one expects similar results under probability weighting and rank-dependent expected utility (or cumulative prospect theory for that matter). Although in theory assuming simply a weighting function w(pi) that is inverse-s-shaped and “very” convex for large gains (see for the associated parametric restrictions) could overturn the rankings of concave and convex exposures stated above, results in [10] suggest that this is not to be expected for the overwhelming majority of the rats that are subject to their experiments.

Choice Data Analysis

Given the four options modelled in the previous section, each of the 20 rats went through 41 final sessions. Denote fA, fF , fR, and fV the relative frequencies (in terms of nose pokes) with which each of the four options Antifragile, Fragile, Robust and Vulnerable, respectively, is chosen over a particular session. In order to represent the rats’ choices, we construct the rotated square in panel (d) of Figure 1 using a linear transformation of the 3-frequency vector, as follows:

In equation (4), the first two rows deliver the two coordinates of interest that are the Total Sentivity to REE (TSREE in short) and One-sided Sentivity to REE (OSREE), while the last row requires that all frequencies sum to one. To derive TSREE and OSREE, it is useful to go through the following steps. Since we are interested in the convexity property of each chosen option, we first measure the frequencies of convex options in the gain domain by summing fA and fV . Similarly, fA + fR measures how frequent choices are convex in the loss domain. It follows that TSREE measures total convexity if it equals 2fA + fV + fR, while OSREE equals fV −fR and measures asymmetric convexity (that in the gain domain minus that in the loss domain). Hence the first two rows in equation (4). Using that fA + fF + fR + fV = 1 - that is, the last row of equation (4) - it follows that TSREE equals the simpler expression 1 + fA − fF . Note that the 3 × 3 matrix in equation (4) is invertible, which implies that all three frequencies can be recovered from the left vector. In addition, that matrix is not unique in the sense that the definitions of TSREE and OSREE that appear above as its first two lines could be moved to any other rows, provided that the identity in the third row and the constraint in the last row are adapted accordingly. This essentially means that given both differences TSREE (= 1 + fA − fF ) and OSREE (= fV − fR), any of the four relative frequencies provides enough information to derive the remaining three, using the constraint that all four frequencies sum up to one. Finally, it follows that in the rotated square in panel (d) of Figure 1, the four vertices have the following coordinates: (0, 0) for Fragile, (0, 2) for Antifragile, (−1, 1) for Robust, and (1, 1) for Vulnerable. TSREE and OSREE are depicted in Figure 2.

TSREE and OSREE are also related to behavioral measures of the share of nosepokes that lead to either exposure or non exposure to REE, as follows: we define Black Swan Avoidance as the share of nosepokes that lead to avoidance of Black Swans - i.e. fA + fR and Jackpot Seeking as the share of nosepokes that lead to be exposed to Jackpots - i.e. fA + fV . It follows that TSREE is by definition the sum of those two shares, while OSREE is the difference between the former and the latter. Formally, if we denote JPS and BSA the shares of nosepokes that lead to be exposed to Jackpots and to avoid Black Swans, this implies that JPS= 0.5(TSREE+OSREE) and BSA= 0.5(TSREE−OSREE). In other words, in the (not rotated) square that appears in panel (d) of Figure 1, the four vertices have the alternative coordinates: (0, 0) for Fragile, (1, 1) for Antifragile, (0, 1) for Robust, and (1, 0) for Vulnerable, in the (JPS,BSA) coordinates. Obviously, both interpretations are equivalent up to a change in coordinates which is a bijection. Note that in the rotated square in panel (d) of Figure 1, lines with a 45-degree slope depict choices with constant BSA so that lines moving north-west represent increasing BSA. Similarly, lines with a −45-degree slope depict choices with constant JPS so that lines moving north-east represent increasing JPS. Median BSA and JP are depicted in Figure 3.

In Figure 4 we report the short term behavioral responsiveness of rat to REE. Behavioral responsiveness is measured by calculating for each rat before/after differences in TSREE and OSREE. This amounts to calculate TSREE and OSREE over the 10 choices preceding a REE and over the 10 choices following the occurrence of each REE. The window of 10 choices before and after is dictated by the configuration of REE in our design. A REE can happen at 11th choice in a sequence (see appendix ). Before/after TSREE and OSREE are averaged by rat over the 41 sessions in the gain domain (Jackpot) and in the loss domain (Black Swan). Table 3 reports the values for the mean before/after differences. In each panel of Figure 4, each rat is depicted by a color-coded dot and the dotted black line represents the 45-degree line. Color-coded dot on the 45-degree line indicate no difference in TSREE or OSREE before and after a REE, i.e. no short term behavioral responsiveness to REE. To visualisation purpose, we computed for each panel a smooth spline regression that estimates a non parametric relationship between before and after TSREE and OSREE. They are plotted as solid black lines in each panel.

We then test the statistical significance of short term behavioral by conducting bootstrap paired-sample mean tests on before/after coordinates (separately) for each type of REE. An observation in the sample is the mean behavioral responsiveness in TSREE or OSREE for a given rat. Observations therefore respect statistical independence. Bootstrap tests lead to the following p-values, under the null hypothesis that the mean difference is zero. Following Jackpots, p = 0.3470 for TSREE and p = 0.0522 for OSREE, which indicates that the null hypothesis is not rejected at 1%. Following Black Swans, however, the null hypothesis is rejected at 1% since p = 0.0045 for TSREE and p = 0.0064 for OSREE. In sum, the mean before/after difference following a REE in the loss domain is significant for both Total and One-sided sensitivities, further confirming the Black Swan avoidance result that we document in this paper.

We also report in Figure 8 how choices by rats after the 41 final sessions compare to choices made in the training step 3, defined in the section on Behavioral Training and Testing. We do so by computing TSREE and OSREE in step 3 sessions and pair coordinates resulting from training with that resulting from the 40 experimental sessions reported in Figure 2. The lines between two color-coded dot connects for each rat its conditioning coordinates with its coordinates resulting from its behavior in the 41 sessions. We report on the right side of the figure the total variation distance between the distribution of nosepokes in the 4 holes in the training and the distribution in the 41 experimental sessions. The interpretation of total variation distance in our setting is simple: it measures, for each rat, the proportion of nosepokes that need to be changed in order to equalize the behavior in the training and the behavior in the 41 final experimental sessions. Median and mean total variation distance are 0.293 and 0.324 respectively, which implies that half of the rats changed more than 30% of their choices in the final experimental sessions compared to the training sessions.

Variation of choices observed in final experimental sessions compared to that observed in training sessions (step 3 in Material and Methods section on Experimental Method Details); panel (a) shows variations of Total Sensitivity to REE (y-axis) and of One-sided Sensitivity to REE (x-axis), that is, dark color-coded dots replicate averages over the 40 sessions for each of the 20 rats - see panel (c) in Figure 2 - and are paired with averages over training sessions for the same rat, depicted with light color-coded dots connected through lines with corresponding dark color-coded dots; panel (b) shows total variation distance, that is, the proportion of total nose pokes in training sessions that need to be changed to replicate nosepokes in final sessions

Finally, in Figure S1 of the Supplementary Material, we report what we label “convexity premiums”, which are counterfactual situations that we construct as follows. Given the rats’ choices over the 41 sessions, we recall all random sequences that have been used to generate gains and losses for each rat. We next compute the outcome that would have been obtained, had the rat chosen convex options in the gain domain (that is, Antifragile or Vulnerable) and in the loss domain (that is, Antifragile or Robust) all the time. The right and left panels in Figure S1 depict the normalized convexity premiums thus computed, for each rat represented in rows, with convex exposures yielding more pellets and implying less waiting in terms of seconds.

Augmented Q-Learning Model Estimation and Simulation

Parameter optimization is carried out for each rat over the 41 sessions, using observed choices and outcomes. We estimate by maximum likelihood two series of augmented Q-Learning models. The first series integrates REE in Q subvalues whereas the second series does not. Each series consists of a first baseline model without specific forgetting rates and decision weights attached to REE. Two partial models integrate either specific forgetting parameters or decisions weights on the presence of REE. The fourth model integrate both.

In total, we estimate eight nested models for each rat. Formally, we estimate four different models: a baseline model (Model 1) without specific forgetting parameters (, with k ∈ {g, l}) and with zero decision weights of REE (γg = γl = 0). Model 2 introduces specific forgetting parameters (, with k ∈ {g, l}) and Model 3 introduces decision weights of REE (γg and γl≠0). Model 4 introduces both.

To ensure convergence and avoid local maxima, each parameter in baseline models was initialized with two different points of its parameter space. This makes in total 42 = 16 combinations of initial parameter values. Each of the 16 models is estimated using an automatized two-step procedure. The fist step consists of an unconstrained downhill simplex method in [29], which does not make use of first derivatives. Estimated parameters are then used as initial parameters in the second step of the estimation procedure, a quasi-Newton method in [16]. Parameter estimates from baseline models are then used as initial parameters for subsequent augmented models. Estimation is carried out using the above two-step procedure to ensure that we convergence is reached.

Once the eight models are estimated, we then compare them using both Akaike information criterion (AIC) and Bayesien Information criterion (BIC), that penalizes the use of additional parameters, so as to select which model configuration best fits observed behavior for each rat (see Supplementary Material for detailed results per rat). We check whether the selected model properly estimates sensitivities to REE by comparing estimated sensitivities to observed sensitivities to REE. This is done by computing Total Sensitivity to REE and One-sided Sensitivity to REE based on estimated choice frequencies from the extended Q-Learning model over the 41 sessions. Fitted sensitivities are reported bottom right panel in Figure 5. Finally, we checked that selected models were able to reproduce experimental data by running simulations for each phenotype observed in the sample of animals (as in [32]). Simulations are conducted over the artificially generated original 41 sessions, using individual parameter estimates for pink and red rats, median parameter estimates for blue and green rats and estimated Q subvalues as prior Qs. For each simulation, we then computed Total Sensitivity to REE and One-sided Sensitivity to REE. Simulations are reported in bottom right panel of Figure 5. See Supplementary Material for more details on the estimation results and for an exploration in related outcome range-adaption models.

eLife Assessment

This study represents an important contribution to the study of decision-making under risk, bringing an interdisciplinary approach spanning economic theory, behavioral neuroscience, and computational modeling to test how choice preference is influenced by rare and extreme events. The authors present evidence that rats are indeed sensitive to these rare and extreme events despite their infrequent occurrence, driven primarily by an almost complete avoidance of "Black Swans" - rare and extreme losses. The evidence for specific sensitivity to rare and extreme events however remains incomplete, owing in part to the difficulty of isolating the effect of these events beyond that arising from risk preferences more generally in both task design and in the computational modeling of the choice behavior. Given the approach here brings a relatively novel perspective, with a more detailed treatment of these confounds this paper will be of broad interest to those seeking to understand animal behavior through the lens of economic choice.

Reviewer #1 (Public Review):

Summary:

In this manuscript, the authors investigate the impact of rare and extreme events on rodents' decision-making under risk, in gain and loss contexts. They describe the behavior of 20 rats performing a four-armed bandit task, where probabilistic gains (sugar pellets) and losses (time-out punishments) can - in some arms - incorporate extremely large - but rare - outcomes. They report that most rats are sensitive to rare and extreme outcomes despite their infrequent occurrence, and that this sensitivity is primarily driven by extreme loss events which they try to avoid, rather than extreme gains that they seek to obtain.

They finally propose a modification of standard reinforcement-learning, which features a specific sensitivity to rare and extreme outcomes and can account for the observed behavior.

Strengths:

The manuscript really taps into a surprisingly neglected but very relevant aspect of decision-making: the effect of rare and extreme events (REE). The authors have developed an experimental setup that seemingly allows investigation of this aspect, which is not trivial given the idiosyncratic properties of rare and extreme events.

The parameters of the experimental setup seem also to be well thought off: basically, in the absence of REE, some options are objectively better than others (because, in expectation, they overall deliver more food, or minimize time-out punishments), but this ordering reverses if REE are taken into account. This allows for a clean test of the integration of REE in the rodent's decision-making model.

The data is presented and analyzed in a very descriptive but exhaustive and transparent way, down to the description of individual rodent's behavior.

Weaknesses:

While the description and analyses of the behavioral patterns are rigorously done under the economic lens of risky decision-making, the authors' interpretation heavily relies on the assumption that rodents have built the correct model of the task during the training. Extensive details are provided about the training procedure, and the observed behavior at the end of the training, but it remains virtually impossible to disambiguate choices due to imperfect learning to choices made due to intrinsic preferences for risk or REE.

By nature, gains (food pellets) and losses (time-out punishments) are somewhat incommensurable so the interpretation of the asymmetry due to outcome valence is also subject to interpretation. There might be some additional subtleties due e.g. satiety that could come from gaining REE (i.e. the delivery of 80 pellets from the Jackpot).

In its current form, the paper is quite hard to digest. This is naturally the case with interdisciplinary work (here mixing economists and neurobiologists). But I am afraid that with the current frame, the paper is going to miss its target, in terms of audience.

The proposed model seems somewhat disconnected from the behavioral patterns: while the model suggests an effect of REE at the decision stage (i.e. with specific decision weights for those rare events), this formalism seems at odds with the observation that REE (notably in the loss domain) has an impact of subsequent behavior - (Black Swans tend to reinforce Total Sensitivity to REE) which rather suggests an effect at the learning stage.

Discussion:

This study convincingly demonstrates that REEs are processed rather uniquely, which makes sense given their evolutionary relevance. REE has indeed been somewhat neglected in previous research, and this study therefore opens an interesting new front on the fundamental aspects of decision under risk. The authors have devised an original theoretical and empirical framework that will be useful for the community, and the combination of economics analysis and rodent behavior constitutes a thought-provoking ground to think about the nature of risk preferences. The interpretation and mechanistic account of these aspects, as well as their generalizability outside the specific context of this study, remain to be strengthened.

Reviewer #2 (Public Review):

Summary:

This paper attempts to examine how rare, extreme events impact decision-making in rats. The paper used an extensive behavioural study with rats to evaluate how the probability and magnitude of outcomes impact preference. The paper, however, provides limited evidence for the conclusions because the design did not allow for the isolation of the rare, extreme events in choice. There are many confounding factors, including the outcome variance and presence of less-rare, and less-extreme outcomes in the same conditions.

Strengths:

(1) The major strength of the paper is the significant volume of behavioural data with a reasonable sample size of 20 rats.

(2) The paper attempts to examine losses with rats (a notoriously tricky problem with non-human animals) by substituting time-outs as a proxy for losses. This allows for mixed gambles that have both gain and loss possible outcomes.

(3) The paper integrates both a behavioural and a modelling approach to get at the factors that drive decision-making.

(4) The paper takes seriously the question of what it means for an event to be rare, pushing to less frequent outcomes than usually used with non-human animals.

Weaknesses:

(1) The primary issue with this work is that the primary experimental manipulation fails to isolate the rare, extreme events in choice. As I understand the task, in all the conditions with a rare extreme event (e.g., 80 pellets with probability epsilon), there is also a less-rare, less-extreme event (e.g., 12 pellets with probability 5). In addition, the variance differs between the two conditions. So, any impact attributable to the rare, extreme event could be due to the less rare event or due difference in the variance. The design does not support the conclusions. Finally, by deliberately confounding rarity and extremity, the design does not allow for assessing the impact of either aspect.

(2) The RL-modelling work also fails to show a specific impact of the rare extreme event. As best as I can understand Eq 2, the model provides a free parameter that adds a bonus to the value of either the two options with high-variance gains (A and V in the paper) or to the two options with high-variance losses (F and V in the paper). This parameter only depends on whether this option could have possibly yielded the rare, extreme outcome (i.e., based on the generative probability) and was not connected to its actual appearance. That makes it a free parameter that just bumps up (or down) the probability of selecting a pair of options. In the case of the "black swan" or high-variance loss conditions, this seems very much like a loss aversion parameter, but an additive one instead of a multiplicative one.

(3) The paper presented the methods and results with lots of neologisms and fairly obscure jargon (e.g., fragility, total REE sensitivity). That made it very hard to decipher exactly what was done and what was found. For example, on p. 4, the use of concave and convex was very hard to decipher; the text even has to repeat itself 3 times (i.e., "to repeat" and "in other words") and is still not clear. It would be much clearer (and probably accurate) to say that the options varied along the variance dimension, separately for gains and losses. Option A was low-variance gains and losses. Option B was low-variance losses and high-variance gains. Option C was high-variance losses and low-variance gains, and Option D was high-variance losses and gains. That tells much more clearly what the animals experienced without the reader having to master a set of new terminologies around fragility and robustness, which brings a set of theoretical assumptions unnecessarily into the description of the experimental design. In terms of results, "Black Swan" avoidance is more simply known as risk aversion for losses.

(4) Were the probabilities shuffled or truly random (seem to be fixed sequences, so neither)? What were the experienced probabilities? Given the fixed sequences, these experienced ("ex-post") probabilities, could differ tremendously from the scheduled ("ex ante") probabilities. It's quite possible that an animal never experienced the rare, extreme event for a specific option. It's even possible (if they only picked it on the 10th/60th choices by chance), that they only ever experienced that rare extreme event. This cannot be known given the information provided. The Supplemental info on p.55 only gives gross overall numbers but does not indicate what the rats experienced for each choice/option-which is what matters here. A simple table that indicates for each of the 4 options, how often they were selected, and how often the animals experienced each of the 6-8 possible outcome would make it much clearer how closely the experience matched the planned outcomes. In addition, by restricting the rare outcome to either the 10th or 60th activations in a session, these are not random. Did the animals learn this association?

(5) The choice data are only presented in an overprocessed fashion with a sum and a difference (in both figures and tables). The basic datum (probability/frequency of selecting each of the 4 options) is not provided directly, even if it can theoretically be inferred from the sum and the difference. To understand what the rats actually do, we first need to see how often they select each option, without these transformations.

(6) There is insufficient detail provided on the inferential statistical tests (e.g., no degrees of freedom or effect sizes), and only limited information on exactly what tests were run and how (bootstrapping, but little detail). Without code or data (only summary information is provided in the supplement), this is difficult to evaluate. In addition, the studies seem not to be pre-registered in any way, leaving many researchers with degrees of freedom. Were any alternative analysis pipelines attempted? Similarly, there were many sub-groupings of the animals, and then comparisons between them - were these post-hoc?

(7) On p. 17, there is an attempt to look at the impact of a rare, extreme event by plotting a measure of preference for the 10 trials before/after the rare, extreme event. In the human literature, the main impact of experiencing a rare, extreme event is what is known as the wavy recency effect (See Plonsky et al. 2015 in Psych Review for example). What this means is that there tends to be some immediate negative recency (e.g., avoiding a rare gain) followed by positive recency (e.g., chasing the rare gain). Using a 10-trial window would thus obscure any impact of this rare, extreme event. An analysis that looks at a time course trial-by-trial could reveal any impact.

(8) As I understood the method (p. 31), the assignment of options to physical locations was not random or counterbalanced, but deliberately biased to have one of the options in the preferred location. This would seem to create a bias towards a particular option and a bias away from the other options, which confounds the preference data in subsequent analyses.

(9) Are delays really losses? This is a big assumption. Magnitude and delay are different aspects of experience, which are not necessarily commensurable and can be manipulated independently. And, for the model, how were these delays transformed into outcomes for the model? Eq 1 skips over that. Is there an assumption of linearity? In addition, I was not wholly clear if the delays meant fewer trials in a session or if the delays merely extended the session and meant longer delays until the next choice period.

(10) The paper does not sufficiently accurately represent the existing literature on human risky decision-making (with and without rare events). Here are a few examples of misrepresented and/or missing literature:

-Most studies on decision-making do not only rely on p > 10% (as per p. 2). Maybe that is true with animals, but not a fair statement generally. Some do, and some don't. There is substantial literature looking at rarer events in both descriptions (most famously with Kahneman & Tversky's work), but also in experience (which is alluded to in reference 19). That reference is not only about the situation when choices are not repeated (e.g. the sampling paradigm), but also partial feedback and full-feedback situations.

The literature on learning from rewarding experiences in humans is obliquely referenced but not really incorporated. In short, there are two main findings - firstly people underweight rare events in experience; second, people overweight extreme outcomes in experience (both contrary to description). Some related papers are cited, but their content is not used or incorporated into the logic of the manuscript.

One recent study systematically examined rarity and extremity in human risky decision-making, which seems very relevant here: Mason et al. (2024). Rare and extreme outcomes in risky choice. Psychonomic Bulletin & Review, 31, 1301-1308.

There is a fair bit of research on the human perception of the risk of rare events (including from experience) and important events like climate. One notable paper is Newell et al (2015) in Nature Climate Change.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

In this manuscript, the authors investigate the impact of rare and extreme events on rodents' decisionmaking under risk, in gain and loss contexts. They describe the behavior of 20 rats performing a four-armed bandit task, where probabilistic gains (sugar pellets) and losses (time-out punishments) can - in some arms - incorporate extremely large - but rare - outcomes. They report that most rats are sensitive to rare and extreme outcomes despite their infrequent occurrence, and that this sensitivity is primarily driven by extreme loss events which they try to avoid, rather than extreme gains that they seek to obtain.

They finally propose a modification of standard reinforcement-learning, which features a specific sensitivity to rare and extreme outcomes and can account for the observed behavior.

Strengths:

The manuscript really taps into a surprisingly neglected but very relevant aspect of decision-making: the effect of rare and extreme events (REE). The authors have developed an experimental setup that seemingly allows investigation of this aspect, which is not trivial given the idiosyncratic properties of rare and extreme events.

The parameters of the experimental setup seem also to be well thought off: basically, in the absence of REE, some options are objectively better than others (because, in expectation, they overall deliver more food, or minimize time-out punishments), but this ordering reverses if REE are taken into account. This allows for a clean test of the integration of REE in the rodent's decision-making model.

The data is presented and analyzed in a very descriptive but exhaustive and transparent way, down to the description of individual rodent's behavior.

Weaknesses:

While the description and analyses of the behavioral patterns are rigorously done under the economic lens of risky decision-making, the authors' interpretation heavily relies on the assumption that rodents have built the correct model of the task during the training. Extensive details are provided about the training procedure, and the observed behavior at the end of the training, but it remains virtually impossible to disambiguate choices due to imperfect learning to choices made due to intrinsic preferences for risk or REE.

As detailed in Material and Methods, the animals were progressively overtrained following standard behavioral procedures. During this process, they experienced all available options, including both positive and negative REE. We assume that repeated exposure to these REE supported learning, as would be expected for any event occurring throughout such an extended training phase. The rats ultimately displayed an asymmetric pattern of choices: they consistently avoided the Black Swan, indicating that they had learned its negative consequences, yet they did not systematically seek the Jackpot. If their behavior were driven solely by incomplete learning or by an inherent preference for risk or REE, we would expect to see the opposite pattern systematic Jackpot seeking or inconsistent avoidance of the Black Swan.

By nature, gains (food pellets) and losses (time-out punishments) are somewhat incommensurable so the interpretation of the asymmetry due to outcome valence is also subject to interpretation. There might be some additional subtleties due e.g. satiety that could come from gaining REE (i.e. the delivery of 80 pellets from the Jackpot).

As described in Material and Methods, we used mouse pellets (20 mg) instead of rat pellets (45 mg) to prevent satiety during Jackpot delivery (80 pellets). We also selected gains (sweet pellets) and losses (delays) that we have successfully used in previous rat decision-making paradigms, such as the rat gambling task (Adams et al., 2017; doi: 10.1523/ENEURO.0094-17) and the loss-chasing task (Breysse et al., 2021; doi: 10.1111/ejn.14895). Notably, if the Jackpot induced satiety, one would expect animals to stop seeking it yet this was not systematically observed. Nonetheless, we added a sentence to the Discussion on page 18 of the manuscript to acknowledge that we cannot fully exclude the possibility that satiety contributed to the lack of systematic Jackpot Seeking.

In its current form, the paper is quite hard to digest. This is naturally the case with interdisciplinary work (here mixing economists and neurobiologists). But I am afraid that with the current frame, the paper is going to miss its target, in terms of audience.

We have rewritten entirely and the english was corrected thanks to ChatGPT. We hope that the paper is now easier to digest.

The proposed model seems somewhat disconnected from the behavioral patterns: while the model suggests an effect of REE at the decision stage (i.e. with specific decision weights for those rare events), this formalism seems at odds with the observation that REE (notably in the loss domain) has an impact of subsequent behavior - (Black Swans tend to reinforce Total Sensitivity to REE) which rather suggests an effect at the learning stage.

We agree with the referee that this may appear surprising at first glance. However, we would first like to emphasize that the general model allows REE to influence learning—that is, to contribute to the updating of the Q-subvalues. Moreover, even when REE are incorporated only as decision weights, as is the case for most rats, this does not imply that REE are unimportant during learning. In fact, the model assumes that REE are learned once and for all when they first occur during a trial of the corresponding option. Unreported simulation exercises indicate that a more gradual learning of maximal and minimal values would likely yield similar results.

Second, the Before/After analysis shows that the behavioral response to Black Swans is locally small in terms of both total and one-sided sensitivities. This suggests that such effects are likely too subtle to be captured by this class of models for most rats. We have added this clarification to the revised version (page 17).

Discussion:

This study convincingly demonstrates that REEs are processed rather uniquely, which makes sense given their evolutionary relevance. REE has indeed been somewhat neglected in previous research, and this study therefore opens an interesting new front on the fundamental aspects of decision under risk. The authors have devised an original theoretical and empirical framework that will be useful for the community, and the combination of economics analysis and rodent behavior constitutes a thoughtprovoking ground to think about the nature of risk preferences. The interpretation and mechanistic account of these aspects, as well as their generalizability outside the specific context of this study, remain to be strengthened.

We have modified the discussion to further insist on the translational aspect of the study and its interest for various populations (page 22). We hope that the generalizability is now strengthened.

Reviewer #2 (Public Review):

Summary:

This paper attempts to examine how rare, extreme events impact decision-making in rats. The paper used an extensive behavioural study with rats to evaluate how the probability and magnitude of outcomes impact preference. The paper, however, provides limited evidence for the conclusions because the design did not allow for the isolation of the rare, extreme events in choice. There are many confounding factors, including the outcome variance and presence of less-rare, and less-extreme outcomes in the same conditions.

Strengths:

(1) The major strength of the paper is the significant volume of behavioural data with a reasonable sample size of 20 rats.

(2) The paper attempts to examine losses with rats (a notoriously tricky problem with non-human animals) by substituting time-outs as a proxy for losses. This allows for mixed gambles that have both gain and loss possible outcomes.

(3) The paper integrates both a behavioural and a modelling approach to get at the factors that drive decision-making.

(4) The paper takes seriously the question of what it means for an event to be rare, pushing to less frequent outcomes than usually used with non-human animals.

Weaknesses:

(1) The primary issue with this work is that the primary experimental manipulation fails to isolate the rare, extreme events in choice. As I understand the task, in all the conditions with a rare extreme event (e.g., 80 pellets with probability epsilon), there is also a less-rare, less-extreme event (e.g., 12 pellets with probability 5). In addition, the variance differs between the two conditions. So, any impact attributable to the rare, extreme event could be due to the less rare event or due difference in the variance. The design does not support the conclusions. Finally, by deliberately confounding rarity and extremity, the design does not allow for assessing the impact of either aspect.

We agree with the referee that both the REE and the rare (≈10% frequency) but non-extreme outcomes are present in the relevant options. However, the rare but non-extreme reward is not large enough to make the convex option attractive and to shift choice away from the concave option. In other words, unlike REE, these outcomes do not reverse stochastic dominance in our design (as noted in Material and Methods). We have explored modified designs for human subjects in which the rare but non-extreme outcomes are removed. Preliminary results indicate that the behavioral phenotypes observed in rats also emerge in humans under these modified conditions, suggesting that REE are the primary drivers. We have added a statement to the Discussion (page 22) to clarify this point.

We elaborate further in our response to point (3) below on why analyses based solely on variance are insufficient when dealing with REE. To clarify the role of rare and extreme outcomes in distinguishing convex from concave options, we provide two new columns to Table 2 in the Materials and Methods, in our reply to point (3).

Finally, although a detailed analysis of rare but non-extreme outcomes lies outside the scope of this paper, the symmetric treatment of extreme and frequent outcomes can be addressed straightforwardly using strong First-Order Stochastic Dominance. Classical decision-theoretic approaches indeed satisfy this property.

(2) The RL-modelling work also fails to show a specific impact of the rare extreme event. As best as I can understand Eq 2, the model provides a free parameter that adds a bonus to the value of either the two options with high-variance gains (A and V in the paper) or to the two options with high-variance losses (F and V in the paper). This parameter only depends on whether this option could have possibly yielded the rare, extreme outcome (i.e., based on the generative probability) and was not connected to its actual appearance. That makes it a free parameter that just bumps up (or down) the probability of selecting a pair of options. In the case of the "black swan" or high-variance loss conditions, this seems very much like a loss aversion parameter, but an additive one instead of a multiplicative one.

We agree with the referee that the additional parameters, compared to more standard Q-learning models, specifically capture the fact that some options deliver REE while others do not. In our estimation procedure, these parameters become nonzero as soon as REE are observed for the first time for a given option. Therefore, the first step is to estimate a baseline nested model in which REEs contribute only at the learning stage (i.e., they affect the updating of Q-subvalues), while the additional parameters are constrained to zero. The next step is to compare alternative models against this baseline, allowing REEs to enter through the additional parameters. In this respect, our specification is parsimonious, especially given that very little is known about REEs in computational neuroscience. More structural modeling is certainly a promising direction for future research, and this paper constitutes a first step toward that goal.

We provide the BIC, in addition to the AIC, to account for the presence of additional parameters in model selection and to ensure that the observed improvement in fit is not merely driven by their inclusion.

Unlike most of the existing literature, our results extend the notion of loss aversion to extreme losses. The negative decision weight on options yielding the Black Swan can be interpreted as a differential treatment of negative REE, an issue we discuss extensively in the Discussion (page 20).

(3) The paper presented the methods and results with lots of neologisms and fairly obscure jargon (e.g., fragility, total REE sensitivity). That made it very hard to decipher exactly what was done and what was found. For example, on p. 4, the use of concave and convex was very hard to decipher; the text even has to repeat itself 3 times (i.e., "to repeat" and "in other words") and is still not clear. It would be much clearer (and probably accurate) to say that the options varied along the variance dimension, separately for gains and losses. Option A was low-variance gains and losses. Option B was low-variance losses and high-variance gains. Option C was high-variance losses and low-variance gains, and Option D was high-variance losses and gains. That tells much more clearly what the animals experienced without the reader having to master a set of new terminologies around fragility and robustness, which brings a set of theoretical assumptions unnecessarily into the description of the experimental design. In terms of results, "Black Swan" avoidance is more simply known as risk aversion for losses.

Because our experimental design focuses on REE, outcomes cannot be summarized only by their variance. This is well known from the large literature on so-called fat-tailed statistical distributions. Unlike the Normal distribution that is entirely characterized by its expected value and variance, fat-tailed distributions have nonzero kurtosis. This implies that a fat-tailed distribution (e.g. exponential) with the same expected value and variance as the Normal differs importantly by possessing extreme values that are much more likely in terms of frequency. To illustrate, if the distribution of pellets was assumed to be Normal with expected value set at 3.89 and variance set at 9.37 as for the convex option, the probability of getting 80 pellets would be about 2.10-16, practically zero. In contrast, this probability is smaller than, but close to 1% in our design.

In Material and Methods, we clearly explain how our novel approach in terms of convexity relates to the moments of the reward distributions, including but not limited to the variance. To clarify further, we provide two new tables (Author response table 2 and Author response table 3) to be compared to Table 2 of the manuscript in which we report the first four moments (mean, standard deviation, skewness and kurtosis) of the full concave and convex gain distributions, reproduced for convenience

In Author response table 2 we report the first four moments when REE are truncated. Comparing convex and concave gains shows that the convex option has a smaller but still close mean compared to the concave option. In contrast, the former has larger variance, skewness and kurtosis compared to the latter. Therefore, interpreting choosing the convex option as reflecting “preference” for variance is at best incomplete.

Author response table 2 further shows that REE alone goes a long way towards explaining the differences between convex and concave options in terms of the first four moments: removing the rare and extreme value results in the concave option having now a larger mean, while the convex option still has larger variance, skewness, and kurtosis but by a smaller margin.

In Author response table 3 we report the first four moments when both RE and REE are truncated, which shows that the convex and concave options differ only with respect to their mean (which is here also larger for concave).

In addition, our focus on REE implies that we go beyond mean-variance preferences that apply mostly to Gaussian distributions. It is not clear theoretically what type of utility functions would reflect preferences that combine a taste for variance, skewness and kurtosis, even though all those moments affect expected utility. See for example Phelps, C.E. “A user’s guide to economic utility functions”. J Risk Uncertain 69, 235–280 (2024) for a recent overview (on page 242, Phelps states that “In situations where risk is not normally distributed, it is ill-advised to ignore statistical parameters beyond variance, unless the deviations from normality are relatively small”).

More importantly, our proposed measure of the convexity of the reward distributions, the Jensen gap, further reveals how even restricting the analysis to the first four moments is incomplete in the sense that it fails to characterize the difference between options: the fifth moment of the concave contributes more the Jensen gap than even kurtosis, while one needs to look at much higher moments to find significant contributions to the Jensen gap for the convex option. In that sense, there is no reason to restrict the analysis to variance, and even to skewness and kurtosis, to compare options, in general and in our particular setup as well. Note that introducing REE would result in convex distributions even in simplified designs, e.g. with 3-value support. Studying REE implies the need to look beyond variance, and our proposal is to use the Jensen gap as a measure of convexity. In the Material and Methods section of the paper, we did not develop an in depth analysis of Jensen gap so as to spare the reader confronted with an already rather technical paper.

We thank the referee for raising the issue of whether variance is a simpler explanation of our results. To keep the main text as short as possible, we chose to refrain from adding technical complexity. We hope we made clear in our reply that the analysis cannot be restricted to variance when studying REE. We believe that Jensen gap is a useful notion in this regard. As our replies will be made publicly available, we chose not to integrate the above discussion in the main text.

(4) Were the probabilities shuffled or truly random (seem to be fixed sequences, so neither)? What were the experienced probabilities? Given the fixed sequences, these experienced ("ex-post") probabilities, could differ tremendously from the scheduled ("ex ante") probabilities. It's quite possible that an animal never experienced the rare, extreme event for a specific option. It's even possible (if they only picked it on the 10th/60th choices by chance), that they only ever experienced that rare extreme event. This cannot be known given the information provided. The Supplemental info on p.55 only gives gross overall numbers but does not indicate what the rats experienced for each choice/option-which is what matters here. A simple table that indicates for each of the 4 options, how often they were selected, and how often the animals experienced each of the 6-8 possible outcome would make it much clearer how closely the experience matched the planned outcomes. In addition, by restricting the rare outcome to either the 10th or 60th activations in a session, these are not random. Did the animals learn this association?

Probabilities are not random and a limited number of fixed sequences has been used, as stated in Material and Methods. We have chosen sequences that satisfy our assumptions about ex-post stochastic dominance reversal of convex over concave options when REE are added. We have added in Table S4 the choice frequencies for all four options. If the animals had learnt the 10th and 60th activation, they would exhibit a strategy in their choice that would tend to be more optimized than what is observed. For example, the options offering the possibility to obtain the Jackpot are not optimal in terms of gains for the frequent events, therefore the animals should tend to select these options only around the 10th and 60th choice. Most of their other choices should favor the options delivering the larger gains in the frequent domain. This is not what is observed. We have added this important point in the discussion (page 18).

(5) The choice data are only presented in an overprocessed fashion with a sum and a difference (in both figures and tables). The basic datum (probability/frequency of selecting each of the 4 options) is not provided directly, even if it can theoretically be inferred from the sum and the difference. To understand what the rats actually do, we first need to see how often they select each option, without these transformations.

As described in Material and Methods, the 4 options are combinations of 2 convex and concave sub-options for gains and losses, which is why our analysis of the behavioral data focuses on convexityrelated total and one-sided sensitivities to REE. The third dimension needed to fully characterize rats’ behavior is simply 1−ffFF, the fraction of non-Fragile choices. In addition, we also provide in Table S4 of the Supplementary Material an alternative interpretation in terms of Black Swan Avoidance and Jackpot Seeking. We have added in Table S4 the choice frequencies for all four options. Finally, all the raw data will be made available with open access and no access codes.

(6) There is insufficient detail provided on the inferential statistical tests (e.g., no degrees of freedom or effect sizes), and only limited information on exactly what tests were run and how (bootstrapping, but little detail). Without code or data (only summary information is provided in the supplement), this is difficult to evaluate. In addition, the studies seem not to be pre-registered in any way, leaving many researchers with degrees of freedom. Were any alternative analysis pipelines attempted? Similarly, there were many sub-groupings of the animals, and then comparisons between them - were these post-hoc?

We understand the concern of the referee for pre-registration of the referee, as an epistemic safeguard to make empirical claims more falsifiable, more transparent, and less dependent on post hoc rationalization. But the contemporary push for preregistration is often presented as an “epistemic improvement,” but in practice it functions largely as a norm of moral regulation, not a scientific necessity. The rhetoric is moralistic: preregistered research is “clean,” “transparent,” “credible,” while non-preregistered work is viewed with suspicion—even when the methodology is sound. This language is not epistemologically neutral; it enforces ought to be done, irrespective of the diversity of legitimate scientific practices.

From a philosophy of science perspective, this is historically and conceptually problematic. Scientific progress has never followed a uniform, rule-based method. As e.g. Feyerabend has argued, major discoveries have emerged precisely because researchers were not bound by predetermined plans: they followed anomalies, improvised, reinterpreted data, and revised methods and hypotheses in light of new evidence — practices that a rigid preregistration ethos can suppress and that are not aligned with how genuine discovery often occurs.

Even from a statistical standpoint, preregistration is far from a panacea. It reduces some degrees of freedom (mainly in confirmatory statistics), but it does not eliminate flexibility; researchers can still choose models, transformations, exclusion rules, stopping rules, etc. And more importantly: reducing flexibility is not inherently epistemically virtuous. Flexibility is often necessary to understand data properly—especially in new paradigms or first-of-their-kind experiments, which is the case for this study. Science needs exploration, opportunism, and theoretical plasticity. Preregistration is compatible with these only if it is treated as one optional tool among many—not as a universal evaluative standard.

As the referee pointed out, this study “taps into a surprisingly neglected but very relevant aspect of decision-making.” Our work is therefore mainly exploratory: the experimental paradigm reveals new behavioral patterns in how rats cope with rare and extreme events, and much of our analysis is necessarily descriptive. We conduct formal inference only where it is methodologically appropriate — the short-term behavioral response to rare events (for which we now provide more details in the Material & methods section p.35) and the estimation of augmented Q-learning models, which follow a standard econometric approach (documented in the Material & Method section–see also our response to recommendation 4). These inferential results support the descriptive patterns that motivate this new line of research.

(7) On p. 17, there is an attempt to look at the impact of a rare, extreme event by plotting a measure of preference for the 10 trials before/after the rare, extreme event. In the human literature, the main impact of experiencing a rare, extreme event is what is known as the wavy recency effect (See Plonsky et al. 2015 in Psych Review for example). What this means is that there tends to be some immediate negative recency (e.g., avoiding a rare gain) followed by positive recency (e.g., chasing the rare gain). Using a 10-trial window would thus obscure any impact of this rare, extreme event. An analysis that looks at a time course trial-by-trial could reveal any impact.

We thank the referee for drawing our attention to the wavy recency effect documented in human experiments. We have added the corresponding reference in the Discussion (page 20). Regarding rats, the Before/After analysis reported in the paper suggests that there is no sizeable immediate recency effect for Jackpots. Even for Black Swans, the immediate recency effect we report remains modest when using a 10-trial window, and the analysis of the choice immediately following a REE does not show evidence of immediate negative recency. This casts doubt on the presence of such an effect in rats.

(8) As I understood the method (p. 31), the assignment of options to physical locations was not random or counterbalanced, but deliberately biased to have one of the options in the preferred location. This would seem to create a bias towards a particular option and a bias away from the other options, which confounds the preference data in subsequent analyses.

We agree that the design incorporated an intentional bias toward the anti-fragile option as a proof of concept. Nevertheless, Figure 8 demonstrates that animals substantially altered their choices between training and final testing, with a median change of approximately 35% across sessions. This indicates that behavior was driven by the structure of possible outcomes rather than by a stereotyped location-based preference.

(9) Are delays really losses? This is a big assumption. Magnitude and delay are different aspects of experience, which are not necessarily commensurable and can be manipulated independently. And, for the model, how were these delays transformed into outcomes for the model? Eq 1 skips over that. Is there an assumption of linearity? In addition, I was not wholly clear if the delays meant fewer trials in a session or if the delays merely extended the session and meant longer delays until the next choice period.

Consistent with established rodent decision-making paradigms (Adams et al., 2017 doi: 10.1523/ENEURO.0094-17; Breysse et al., 2021 doi: 10.1111/ejn.14895), we employed sweet pellets as gains and imposed delays as losses. Delays are operationalized as losses because they preclude the animal from engaging in reward-generating behavior; thus, increasing the delay duration proportionally increases the magnitude of the opportunity cost.

(10) The paper does not sufficiently accurately represent the existing literature on human risky decision-making (with and without rare events). Here are a few examples of misrepresented and/or missing literature:

Most studies on decision-making do not only rely on p > 10% (as per p. 2). Maybe that is true with animals, but not a fair statement generally. Some do, and some don't. There is substantial literature looking at rarer events in both descriptions (most famously with Kahneman & Tversky's work), but also in experience (which is alluded to in reference 19). That reference is not only about the situation when choices are not repeated (e.g. the sampling paradigm), but also partial feedback and full-feedback situations.

We have corrected that statement in the main text (page 3) and we thank the referee for pointing this out.

The literature on learning from rewarding experiences in humans is obliquely referenced but not really incorporated. In short, there are two main findings - firstly people underweight rare events in experience; second, people overweight extreme outcomes in experience (both contrary to description). Some related papers are cited, but their content is not used or incorporated into the logic of the manuscript.

One recent study systematically examined rarity and extremity in human risky decision-making, which seems very relevant here: Mason et al. (2024). Rare and extreme outcomes in risky choice. Psychonomic Bulletin & Review, 31, 1301-1308.

There is a fair bit of research on the human perception of the risk of rare events (including from experience) and important events like climate. One notable paper is Newell et al (2015) in Nature Climate Change.

We agree with the referee that the related literature on REE in animal Decision Making is scant and that it is more developed in humans. We thank the referee for pointing at Mason et al. (2024), who clarify where the literature on humans stands and why combining rarity and extremity, as we also do, is important and highly relevant. We have added a new statement and references in the Introduction and Discussion (pages 3, 20, 22).

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

(1) As said above, I think the manuscript would really benefit from a rewriting, to replace some technical terms with more readable ones, and maybe rebalance the focus from the current focus on the framework (heavily loaded with economics concepts, which will be hard to digest for the eLife readership) to a higher weight on information that is critical to understand and interpret the behavior (e.g. information about training & training behavior, etc.).

We have revised the entire manuscript to improve readability and have clarified in the main text: (1) why convexity of exposures to REE could, beyond variance, be useful for experiments in other settings that our own; (2) why the associated notion of antifragility may be applicable to other settings and therefore of broader interest; (3) what was done in the training sessions compared to the final sessions.

(2) From Figure 8, it seems that rodent behavior is more clustered after the training (i.e. before the sessions) than after the sessions. Could that be a sign of imperfect learning?

Figure 8 mostly suggests that there is some flexibility in the choices made and that the intended initial bias towards the antifragile choice in the design of the task could be over ridden by the rats.

(3) The modelling section seems incomplete. I think the authors want to tease apart where REE enters the model and should propose an alternative where REE affects the learning rather than the decision.

In fact, the general model allows REE to have an effect at the learning stage only (i.e. to contribute to the updating of the Q subvalues), when the specific decision weights attached to options delivering REE are both zero. However, our analysis shows that such a model is rejected by the behavioral data for all rats. We have clarified this point in the revised version.

(4) Also, parameter and model recovery exercises seem mandatory (Wilson & Collins, 2019).

We thank the referee for highlighting this valuable reference in computational modeling, particularly in the context of model identification and estimation in computational biology. In the present research, we adopted an econometric perspective on model identification—especially with regard to the integration of Q-values for gains and losses. The softmax choice function is formally equivalent to a multinomial logit model, and as is well known in econometrics, identification in such models presents non-trivial challenges. The standard approach in classical Q-learning is to multiply the Q-value by an inverse temperature parameter (also known as a precision parameter in random utility models). When extending the model to include separate Q-values for gains and losses, specifying the model in an identifiable way becomes more complex.

To address this issue, we considered several alternative model specifications and conducted grid-based estimation of starting parameter values. This approach allowed us to examine the shape of the loglikelihood function and assess whether the parameters are globally identified, rather than only identifiable up to a linear combination. We found that the most parsimonious and empirically identified specification in our experimental paradigm is one in which Q-values for gains and losses are summed, each weighted by distinct decision weights (see our Equation 2 in the paper).

The inclusion of decision weights for REE for each option (Equation 2) is then structurally equivalent to introducing constant terms in a logit model. The identification of these parameters follows standard econometric results on discrete choice models (e.g., Davidson & MacKinnon, 2003): since we model choices among four options, three free parameters can be estimated, leaving one degree of freedom in the specification. As mentioned in the "Modelling and Statistical Analysis" section, we further guarded against the presence of local maxima by applying a two-step estimation procedure, combining two optimization algorithms with multiple sets of starting values for the baseline model (i.e., the model without decision weights for REE). We also tested the addition of a global optimization method— simulated annealing—but found that it did not significantly improve upon our two-step procedure. This is not surprising, as our preliminary investigation of model identification, based on grid searches over starting parameter values, confirmed that all parameters were identified in our simple specification. Our intuition is that simulated annealing may yield different estimates than gradientbased methods primarily in cases where the model is not theoretically identified—suggesting that the need for such global optimization techniques can be indicative of underlying identification issues in Qlearning models.

Regarding model comparison, we have used penalized information criteria to account for additional parameters. Although we do not report confusion or inversion matrices for our nested models, we verified that the estimated models replicate observed behaviors across all phenotypes, as shown in the main text (see bottom left panel of Figure 5 for the Total and One-Sided sensitivities). Most importantly, we conducted 100 additional simulations of 40 artificial sessions for each phenotype using the “winning” models and the median fitted parameters. These simulated rats—playing the task 100 times over 40 sessions—offer strong evidence that the selected models are valid: they quantitatively capture the behavior of all phenotypes in terms of our key metrics, Total and One-Sided sensitivities (see bottom right panel of Figure 5).

Taken together, this methodical econometric approach to model specification and estimation gives us strong confidence in the identification and robustness of our model. Overall, while Wilson & Collins (2019) provide an interesting framework for model estimation in computational biology, we believe that a more formal theoretical analysis of model identification in Q-learning models would be a valuable addition to the field—though it lies beyond the scope of the present work. In our view, computational biologists should complement simulation-based validation and empirical fit with formal methods for assessing theoretical identifiability, particularly when estimating complex choice models.

Davidson, R. and J.G. MacKinnon (2003) Econometric Theory and Methods. Oxford University Press (New York).

Wilson, R. C., & Collins, A. G. (2019). Ten simple rules for the computational modeling of behavioral data. eLife, 8, e49547. https://doi.org/10.7554/eLife.49547

Reviewer #2 (Recommendations For The Authors):

(1) The paper confuses risk sensitivity and exploration in the opening lines. These are not the same.

What we have in mind here is that uncertainty about outcomes is one of the main drivers of exploration, in the sense that there would be no need to explore in a counterfactual world with deterministic gains and losses. We have modified the opening lines of the paper to better reflect this dimension (page 2).

(2) p. 9. "awfully long" is an unnecessary descriptor. Descriptions of methods should be more factual.

The manuscript has been entirely rewritten.

(3) p. 13. Most points lie on the left of the square (not right?).

We thank the referee for pointing at this typo, that is now corrected in the text (page 8).

(4) p. 13. Last line. "obviously" is patronizing to the readers.

The manuscript has been entirely modified to address related points.

(5) p. 23. The avoidance of black swans by not choosing that option sounds like a hot-stove effect (see Denrell & March, 2001). Is this evidenced here?

To the best of our knowledge, the statement that “people tend to avoid activities they have had a negative experience of, resulting in a negativity bias” (from Jerker Denrell’s website) does not explicitly concern REE. Instead, it appears to refer broadly to reinforcement learning mechanisms driven by negative outcomes, irrespective of their magnitude or frequency. In our task, animals encounter both negative rare events (RE) and negative rare and extreme events (REE; Black Swans). Notably, the task design does not allow rats to completely avoid negative RE unless they cease performing the task altogether—a pattern typically seen in paradigms involving aversive stimuli such as electric foot shocks. The fact that all 20 rats maintained stable performance across the 41 sessions provides evidence against a pronounced hot-stove effect. This point has been incorporated into the revised discussion (page 20).

(6) "menus" is an odd term. Better described as reward schedules?

“Menu” has been replaced by “option” in the main text.

(7) Why are they 20-minute sessions? I thought it was 120 trials per session? And 41 sessions? Or was this only in training?

Each session ended after 20 minutes had elapsed, which led to approximately 120 trials (but not systematically). The choice of 20 minutes was made in order to limit the number of trials to prevent satiety. The total number of sessions ran with all 20 animals for the final testing was 41, an odd number but there was no justification to remove one session from the analysis. The training was much longer and is not included in the 41 sessions.

(8) Really not clear why these Jensen inequalities were relevant or even calculated for these options? How is it relevant to what animals chose or experienced? They seem to be based on the generative probabilities for different options, which is not what happened in reality.

We propose the Jensen gap as a general measure of convexity that relates to all moments of the probability distribution, as described in more detail in our answer to point (3) above. As such, we think it is a characterization of options with stochastic outcomes that could prove useful to other experimenters in alternative settings beyond our own.

(9) Only some summary data in supplemental materials. No open data or code for recreating the experiment or analyzing the data.

The data is available on Github (see page 38) and the code will be available upon request.

Version 2

Abstract

Most behavioral studies in animals investigate risk using outcome probabilities larger than 10%. However, real-world Decision-Making often requires evaluating events that are both extremely unlikely and highly consequential. To address this gap, we developed an experimental and computational framework to quantify how rats detect and adapt to rare (probability < 1%) and extreme (deviation from mean > 10 standard deviations) outcomes (REE). Using a four-armed bandit task, animals chose between options associated with probabilistic rewards (sugar pellets) or punishments (time-out delays). Depending on the animal’s choice, REE can occur or not, allowing us to probe how rats integrate information across both common and fat-tailed event distributions. Across subjects, behavior showed restricted diversification (typically two out of four options) and clear sensitivity to REE, expressed as a systematic avoidance of rare and extreme punishments (“Black Swans”) combined with partial exposure to rare and extreme gains (“Jackpots”). The dominant behavioral phenotype displayed a near-complete suppression of exposure to Black Swans, whereas exposure to Jackpots remained only moderate. This asymmetric sensitivity came at a cost: these animals accepted smaller frequent gains and larger frequent losses to avoid catastrophic outcomes. To account for these behaviors, we implemented an augmented reinforcement learning model in which REE are weighted separately from frequent events. Fitting this model to individual behavioral data captured subjects’ decision patterns far better than standard Q-learning, which systematically failed to reproduce the observed asymmetry. The best-fitting model suggests that the rat brain segregates information from the central outcome distribution and the distribution tails, assigning distinct value weights to REE during action selection. This work provides the first evidence that rodents spontaneously adapt their learning strategies to avoid extreme punishments while partially maintaining exposure to rare and extreme gains, suggesting that frequent outcomes are treated separately from rare and extreme ones through different weights in Decision-Making.

Introduction

Exploration and exploitation are fundamental components of adaptive behavior under uncertainty in both humans and non-human animals (for review see [28]). However, relatively little is known about how these behavioral processes are modulated by outcomes that are both extremely rare and highly consequential - hereafter referred to as Rare and Extreme Events (REE). Although risk processing has been extensively studied across psychology, behavioral economics, and neuroscience, most laboratory decision-making tasks rely on outcome distributions in which event probabilities exceed 10% (e.g. [9], [10], [41], [53]; see also [20], [27], [32] for notable exceptions). This represents a major limitation considering that in natural environments, many species must cope with events that occur far less frequently yet exert disproportionately large negative or positive impacts. Such events have shaped evolutionary trajectories and can even lead to species-wide extinction [19], while in humans they have occasionally driven technological and cultural leaps [29]. Because their probabilities and magnitudes cannot be reliably estimated from experience, REE pose profound challenges for learning and decision-making mechanisms. It remains largely unknown whether, and how, animals - including humans - detect, integrate, or adapt to such events.

Here, we address this gap by studying rats, a species widely used in behavioral and systems neuroscience to investigate the neural substrates of cost-benefit learning and decision-making (e.g. [6], [1], [49], [51], [24], [33], [47], [21], [58]; see [23] for relevance to human decision processes). We developed a novel four-armed bandit task in which rats repeatedly interact with their environment and encounter REE with probabilities < 1%. Critically, these REE produce outcome magnitudes - both gains and losses - that exceed 10 standard deviations relative to the average outcome. Standard rodent analogs of the Iowa Gambling Task typically use lower-magnitude rare events with probabilities ≥ 10% ([4], [1], [37], [57]). The central feature of our task is that it allows us to test whether rats adopt behavioral strategies that minimize exposure to extreme losses (Black Swans) while preserving the possibility of extreme gains (Jackpots) - analogous to “Anti-fragile” strategies described in conceptual work [48]. In contrast, strategies that eliminate Jackpots while maintaining exposure to Black Swans are here termed “Fragile”. Operationally, complete exposure to Jackpots and complete avoidance of Black Swans defines the Anti-fragile o ption, whereas the inverse pattern defines the Fragile o ption. We additionally define the Robust option, as the one excluding all REE and Vulnerable option the one exposing animals to both positive and negative REE. This framework is compatible with evolutionary hypotheses proposing that organisms may be biased toward choice policies that enhance long-term viability under uncertainty.

Our design separates outcomes into three probabilistic domains: (i) “normal events” (NE, ≈ 90% probability), (ii) “rare events” (RE, ≈ 10%), and (iii) REE (< 1%). Reward magnitudes scale nonlinearly with decreasing probability such that convex options produce increasingly larger gains as probability decreases (i.e., exposure to Jackpots) while concave options limit the possible gain within a short range whatever the probability (i.e. avoidance of Jackpots). In contrast, in the negative domain, concave options produce increasingly larger losses (exposure to Black Swans), while convex ones limit the possible losses within a short range whatever the probability. Convexity here refers not to utility curvature but to the accelerating relationship between magnitude and rarity.

Based on stochastic dominance arguments, value-maximizing agents with non-decreasing value functions should prefer concave options in the NE domain (first-order dominance), and continue to do so in the RE domain if they also display risk aversion (second-order dominance). However, in the full environment including the REE, second-order dominance reverses, and convex options become optimal because avoiding Black Swans is increasingly costly when REE fail to materialize. Thus, Anti-fragile strategies are only advantageous if REE are internally represented.

We incorporate both gains (sugar pellets) and losses (time-out punishment) to capture sensitivity across valence domains. We derive two key behavioral metrics: (1) Total Sensitivity to REE, defined as the combined frequency of convex choices in gain and loss domains, indexing whether rats incorporate REE into their strategy; and (2) One-Sided Sensitivity to REE, defined as the difference between convex gain choices and convex loss choices, quantifying in particular asymmetric Jackpot-seeking versus Black Swan avoidance.

Using 20 rats (about 6000 trials per subject over 41 sessions), we report two major behavioral findings. First, rats diversify their behavior across options but typically concentrate on roughly two of the four available strategies. Nineteen of 20 animals show moderate-to-high Total Sensitivity, indicating that most animals systematically combine some exposure to extreme rewards with partial avoidance of extreme losses. Second, 13 of 20 rats demonstrate near-complete avoidance of Black Swans while only partially seeking Jackpots, producing negative One-Sided Sensitivity and consistent choice mixtures of Anti-fragile and Robust options. This suggests differential weighting of rare positive versus rare negative outcomes. Consistent with this interpretation, both Total Sensitivity and Black Swan avoidance increase after rats directly experience a rare extreme loss. To interpret computationally these results, we evaluated a family of augmented Q-learning models that incorporate explicit sensitivity parameters for Jackpots and Black Swans. Model comparison using information criteria identifies an architecture in which NE+RE and REE are weighted separately during action selection, suggesting that rats assign distinct cognitive or neural salience to REE - especially negative ones - when making decisions under uncertainty.

Results

Measures of sensitivity to Rare and Extreme Events in a four-armed bandit task

Rats were tested in an operant chamber equipped with four nose-poke ports, each linked to a specific distribution of reward (sucrose pellets) and punishment (time-out delays) drawn from a fixed schedule across sessions (Fig. 1a). After training, animals completed 41 experimental sessions (20 min/session; ≈ 120 trials/session). As shown in Figure 1b, positive outcomes consisted of sucrose pellets, whereas negative outcomes corresponded to time-out periods during which ports were inactive, preventing pellet acquisition.

Experimental Design.

(a) Schematic of the operant conditioning chamber. Rats performed a self-paced four-option choice task using nose-poke apertures, each corresponding to a distinct option leading to distribution of reward (sucrose pellets) or punishment (time-out delay) at various size and probability. (b) Outcome structure for each option. Rewards and losses follow either convex (green) or concave (red) probability-magnitude functions. Outcomes are grouped into three probability domains: high-probability Normal Events (NE; 20 − 25%, blue), Rare Events (RE; 5%, yellow), and Rare and Extreme Events (REE; < 1%, pink). (c) Functional characterization of the four choice options based on exposure to rare and extreme gains (“Jackpots”) and rare and extreme losses (“Black Swans”). On the horizontal x-axis are reported (in decreasing order moving away from the origin) the ex-ante probabilities that are unknown to the subject, which only observes the outcome that is measured on the vertical y-axis. For convenience, gains appear in the upper-right quadrant while losses appear in the lower-left one. Convex curve are in green while concave are red. Robust: convex losses and concave gains (avoids both REE). Anti-fragile: convex losses and convex gains (avoids Black Swans while retaining access to Jackpots). Fragile: concave losses and concave gains (exposure to Black Swans, avoidance of Jackpots). Vulnerable: concave losses and convex gains (exposed to both Black Swans and Jackpots). (d) Behavioral sensitivity metrics. Total Sensitivity (y-axis) quantifies overall incorporation of REE by summing convex choices in gain and loss domains. One-sided Sensitivity (x-axis) measures asymmetry between REE seeking and REE avoidance. Example profiles illustrate: an Anti-fragile-like pattern (green), a Fragile-like pattern (red), and a mixed Anti-fragile/Robust profile with strong Black Swan avoidance and partial Jackpot seeking (blue). See Material and Methods for computational definitions.

Each choice option was defined by a characteristic mapping between outcome magnitude and event probability. Outcome magnitudes could follow either convex (accelerating; green) or concave (decelerating; red) functions as event probability decreased (Fig. 1b-c). Critically, all options included a spectrum of outcome classes: highly probable Normal Events (NE; ≈ 90%), Rare Events (RE; ≈ 10%), and Rare and Extreme Events (REE; ≈ 1%), which corresponded to exceptionally large rewards (“Jackpots”: 80 pellets) or exceptionally large punishments (“Black Swans”: 240-s delay).

Combining convex or concave gain and loss functions generated four distinct options (Anti-fragile, Robust, Vulnerable, Fragile; Fig. 1c). The Anti-fragile option paired convex gains with convex losses, thereby allowing exposure to Jackpots while avoiding exposure to Black Swans. The Fragile option combined concave gains and concave losses, yielding small plateauing rewards but exposing animals to extreme losses (Black Swans). Robust and Vulnerable options showed complementary patterns: Robust avoided Black Swans but also prevented access to Jackpots, whereas Vulnerable exposed animals to both extreme outcomes.

Because the behavioral structure explicitly manipulated exposure to REE, we quantified how frequently each rat selected options containing convex (versus concave) outcome functions. Two measures were derived (Fig. 1d). Total Sensitivity to REE reflects the overall tendency to select options that allow contact with extreme outcomes (i.e., convex functions in either domain). One-sided Sensitivity quantifies asymmetry between the gain and loss domains - specifically, whether an animal preferentially seeks extreme rewards (Jackpot Seeking) or preferentially avoids extreme punishments (Black Swan Avoidance).

These two dimensions generate a behavioral space in which animals may cluster near extreme phenotypes (e.g., exclusively Anti-fragile or exclusively Fragile) or diversify among several strategies. Points lying between edges reflect mixed strategies. For example, a profile near the Anti-fragile - Fragile axis indicates strong Black Swan Avoidance combined with frequent-but not exclusive - Jackpot seeking.

Importantly, although Normal and Rare Events dominate the experience of the task, the inclusion of REE changes the adaptive landscape: extreme outcomes, despite their low probability, produce large shifts in reinforcement history. Thus, rats that behave as though REE “matter” (high Total Sensitivity) preferentially select options allowing contact with large rewards while minimizing catastrophic time-outs, whereas rats that behave as if REE are irrelevant (low Total Sensitivity) preferentially sample concave options that yield stable but limited outcomes, according to the stochastic dominance of order 1. Together, this framework allows quantification of how individual animals integrate extremely infrequent but high-impact outcomes into their decision strategy, enabling downstream analyses linking these behavioral phenotypes to neural processes.

Main phenotype: strong avoidance of extreme punishments with partial sampling of extreme rewards

We analyzed the behavioral data collected across 41 daily sessions from 20 rats (see Material and Methods). To highlight the dominant features of the behavioral strategies adopted by the animals, Figure 2 plots each rat’s choices within the two-dimensional space defined by Total Sensitivity to REE and One-sided Sensitivity (relative weighting of extreme losses vs extreme rewards). This representation corresponds to the rotated square introduced in Figure 1d, where each vertex corresponds to exclusive selection of one of the four outcome distributions. A first inspection of Figure 2 shows that no animal operates at a single vertex - i.e., no rat commits exclusively to one contingency. Instead, all animals sample from multiple options, producing distributed behavioral profiles rather than categorical strategies. This diversification is qualitatively consistent with what is often observed in naturalistic foraging scenarios in which animals hedge against uncertainty or extreme events ([5], [40], [43], [48]).

Total sensitivity to Rare and Extreme Events (REE) (y-axis) plotted against one-sided sensitivity to REE (x-axis), where rightward movement indicates REE seeking and leftward movement indicates REE avoidance.

For a detailed description of the modeling and statistical analysis methods used, please refer to the Materials and Methods section. (a) the 820 black dots represent data from 41 sessions conducted with a total of 20 rats. (b) individual rat profiles, with each square corresponding to a single rat. The colored dots within each square denote the performance across the 41 sessions for each rat, categorized into two primary sensitivity profiles: blue dots indicate rats that exhibit strong sensitivity to REE and are classified as “Black Swan avoiders”, while green dots indicate rats that are responsive to REE but maintain a neutral stance, situated near the midline of the graph. Additionally, two outlier data points are highlighted in red and pink. (c) color-coded dots, each representing the average sensitivity measures across all 41 sessions per rat (n=20 rats).

Population-level patterns

Panel 2a shows the full dataset (≈ 800 session-points). Most sessions lie on the left half of the behavioral space, corresponding to strategies that rather avoid REE. Within this domain, most sessions fall in the upper quadrant, indicating moderate to high Total Sensitivity - that is, rats systematically adjusted their choices in ways that reflect the presence of rare and extreme outcomes by minimizing exposure to extreme losses (Black Swans). Training history contributing to these patterns is detailed in Material and Methods.

Individual behavioral profiles

Panel 2b displays each individual rat’s pattern across the 41 sessions. Each square corresponds to one rat, and the four vertices correspond to the four possible outcome distributions (A: Antifragile; R: Robust; V: Vulnerable; F: Fragile). With the exception of one animal (rat 17, red), all rats show moderate to high Total Sensitivity. Within this dominant group, two sub-groups emerge (blue and green). A second outlier, rat 19 (purple), is the only animal expressing both very high Total Sensitivity and strong Jackpot Seeking, achieved by alternating mainly between Antifragile and, less frequently, Vulnerable options. The remaining 18 animals fall into two stable phenotypes:

Blue group (13 rats): These animals primarily oscillate between the Robust and Antifragile options. This phenotype combines (i) moderate to high Total Sensitivity with (ii) pronounced avoidance of Black Swans. In other words, these rats preferentially avoid extreme punishments more than they seek extreme rewards.

Green group (5 rats): These rats primarily alternate between the Fragile and Antifragile options. They show moderate to high Total Sensitivity but display no directional bias in One-sided Sensitivity, indicating neither consistent avoidance of extreme losses nor active seeking of extreme rewards. Their choices are symmetric with respect to REE in the gain and loss domains.

Average phenotypes

Figure 2c summarizes each rat’s average metrics across sessions. All but four rats exhibit an average Total Sensitivity above half of the maximum value; three are slightly below one, and one rat (rat 17) sits around 0.5, making it the closest to a Fragile-like pattern, though still deviating from true Fragile behavior. In total, 19 of 20 rats show clear sensitivity to the presence of REE.

Panel 2c also reveals a pronounced dominance of Black Swan Avoidance (negative One-sided Sensitivity) in the majority of animals. Rats 17 and 19 are opposite outliers: rat 17 has low Total Sensitivity, whereas rat 19 shows the highest Total Sensitivity coupled with strong Jackpot Seeking.

Within the 18-rat majority, the two main phenotypes described above are clearly distinguished: the green subgroup shows no One-sided bias, while the blue subgroup shows pronounced Black Swan Avoidance. Quantitative values are provided in Supplementary Material: Behavioral Measures per Rat.

Structure in the relationship between Total and One-sided Sensitivity

A notable feature in panel 2c is a positive relationship between Total Sensitivity and One-sided Sensitivity among the 16 rats with the highest Total Sensitivity. Within this high-sensitivity group, animals that more strongly incorporate REE into their choices tend to show reduced Black Swan Avoidance (i.e., their One-sided Sensitivity becomes less negative). This shift reflects a progressive tilt toward convexity in the gain domain relative to the loss domain. Rat 19 represents an extreme case, approaching the maximal combination of Total Sensitivity and Jackpot Seeking.

This relationship is confirmed statistically: in the full dataset of 20 rats, the correlation between average Total and One-sided Sensitivity is near zero. However, restricting the analysis to the 16 high-sensitivity rats yields a correlation of ≈0.57. Over learning, this relationship strengthens: splitting the 41 sessions into four blocks of 10, the correlation increases from ≈0.17 initially to ≈0.56, ≈0.71, and ≈0.64 in successive epochs. Thus, the structure linking Total and One-sided Sensitivity emerges early and stabilizes over time.

Behavioral measures of reward seeking and punishment avoidance

Figure 3 reports more intuitive behavioral metrics: the proportion of choices that (i) avoid Black Swans and (ii) seek Jackpots (detailed computation in Materials and Methods: Modeling and Statistical Analysis). Median values show a sharp dissociation between the two phenotypes. Approximately half of the blue rats avoid extreme punishments in > 90% of trials, whereas only ≈40% of green rats-and none of the outliers-achieve comparable avoidance.

Behavioral Differences Between Phenotypes in Avoidance and Seeking Behaviors.

This figure illustrates the median behavioral measures of Black Swan Avoidance and Jackpot Seeking across the two phenotypic groups, blue and green. The phenotypes are defined by distinct responses to Rare and Extreme Events (REE), with the blue phenotype exhibiting higher sensitivity to REEs and a tendency towards avoiding Black Swans, while the green phenotype shows less bias and remains closer to neutral behavior. Left Panel: This panel depicts the percentage of nosepokes where rats were exposed (black) or not exposed (light gray) to Black Swans. Increased Black Swan Avoidance is indicated by a higher proportion of non-exposure, suggesting a behavioral preference to avoid aversive outcomes among the blue phenotype compared to green. Right Panel: Conversely, this panel shows the percentage of nosepokes where rats were exposed (black) or not exposed (light gray) to Jackpots. Enhanced Jackpot Seeking is indicated by a higher proportion of exposure, suggesting greater motivation towards reward acquisition in the blue phenotype compared to green. Statistical comparisons between the phenotypes were conducted using Wilcoxon Rank Sum tests. Significant differences are denoted with ⋆ (p < 0.05) and ⋆⋆ (p < 0.01).

By contrast, Jackpot Seeking does not differ between blue and green rats. Wilcoxon paired tests confirm this: blue rats differ significantly between punishment avoidance and reward seeking (p = 0.0002), whereas green rats do not (p = 0.6250). The two outliers diverge in opposite directions: rat 17 seeks almost maximal Jackpot exposure, while rat 19 nearly eliminates it. Exact values are given in the Supplementary Material.

Black Swan outcomes strengthen sensitivity to rare and extreme events, whereas Jackpots do not

After establishing that most animals (19/20) display moderate-to-high Total Sensitivity to REE and robust avoidance of extreme punishments (Black Swans), we next asked how the actual experience of an REE shapes subsequent choice behavior within the same session. Specifically, we compared each rat’s behavioral sensitivity in the 10 nose-pokes preceding an REE with the 10 nose-pokes following it.

Figure 4 summarizes these within-session dynamics. For each rat (color-coded as in Figure 2), panel 4a shows the effects of experiencing a Jackpot, whereas panel 4b shows the effects of encountering a Black Swan. Total Sensitivity is plotted on the left column, and One-Sided Sensitivity (relative weighting of extreme losses vs gains) on the right. The dotted diagonal indicates no change.

Responses of Total Sensitivity to REE and of One-sided Sensitivity to REE, averaged over the 41 final sessions, following Jackpots in panel (a) and Black Swans in panel (b), for each of the 20 rats.

Each dot indicates one rat and dots are color coded to indicate the profile of the animal, as in Figure 2. Black dotted lines materialize no change in sensitivities after exposure to the REE compared to before. Black solid lines represent spline estimates -see Material and Methods.

A clear asymmetry emerges: Jackpots produce minimal change in either metric, with most points lying close to the diagonal. In contrast, Black Swan outcomes shift behavior systematically. Most animals show an increase in Total Sensitivity - i.e., they become more responsive to the structure of rare events-along with a decrease in One-Sided Sensitivity, reflecting a further shift toward avoidance of extreme losses. In both measures, points cluster above (Total Sensitivity) or below (One-Sided Sensitivity) the 45° line, indicating a consistent strengthening of REE-related behavioral weighting after punishment but not after reward.

Thus, rare and severe losses - but not rare gains-acutely potentiate the animals’ sensitivity to REE. This pattern aligns with our overall finding that rats prioritize avoidance of extreme negative outcomes and that negative REE have disproportionate influence on choice updating. Bootstrapped statistics and additional tests reported in Material and Methods (see Table 3) confirm the significance of these effects.

In summary, within-session analyses reveal a strong valence asymmetry: Black Swan events transiently enhance both global sensitivity to REE and the bias to avoid rare losses, whereas Jackpots do not measurably alter behavioral sensitivity. This suggests that rare aversive outcomes exert a privileged influence on decision-making processes in this task, reinforcing the dominant phenotype of heightened vigilance toward extreme negative events.

Q-Learning models require specific REE decision weights to mimic rats’ behavior

We have introduced sensitivity to Black Swans and Jackpots in a new class of augmented Reinforcement Learning models and we have estimated their parameters using observed choices and outcomes for each rat. The selected model ended up with a distinctive feature: it separates normal from rare and extreme outcomes through different weights in the Decision-Making process. Adding such specific sensitivity results in a good fit of the selected model - and simulated behaviors that are close - to behavioral observations, whereas a standard Q-learning model without sensitivity to REE is rejected for almost all rats.

To each of the four available options, indexed below by “o” (from 1 to 4, referring to Antifragile, Fragile, Robust, Vulnerable) in equations (1), we attach at each moment in trial/time t a gain sub-value Qg(o) and a loss sub-value Ql(o) that are updated as follows: (1) Qt+1g(o)−Qtg(o)⏟update of gain sub-value =αg⏟gain learning rate ×[rtg(o)−Qtg(o)]⏟gain prediction error Qt+1l(o)−Qtl(o)⏟update of loss sub-value =αl⏟loss learning rate ×[rtl(o)−Qtl(o)]⏟loss prediction error where the convention that updated values are indexed by t + 1 means right after observing the gain rtg or loss rtl corresponding to the choice made at t. Parameters αg and αl are usually labeled learning rates, indicating the speed at which reward prediction error gets updated into the latest values. Limiting cases occur when setting the learning rate either to zero - its smallest possible value - which means no subvalue updating, or to one - its largest possible value - which means that the subvalue tracks the obtained reward itself. While equations (1) applies for the updating of the chosen option, a similar updating rule applies for the three other options that are not chosen at that moment, by assuming that this happens as if the reward was zero, with different parameters. That is, for options that are not chosen, the updating rules for subvalues are simply Qt+1g(o)=(1−αgf)Qtg(o) and Qt+1l(o)=(1−αlf)Qtl(o), where the forgetting rates αgf and αlf are assumed to be bounded between zero and one.

The novel feature of our augmented Q-Learning model, besides integrating gains and losses, is that specific weights are attached to options that may produce REE. More specifically, we model the value of each option o as: (2) Vt(o)⏟value of option=λgQtg(o)+λlQtl(o)⏟decision weight on sub-values+γg1JP(o)+γl1BS(o)⏟decision weight on REE In equation (2), the value of each option Vo is the sum of two terms. The first is the decision weight attached that sums up the gain and loss subvalues, each weighted by parameters λg and λl that may reflect a differential effect. Importantly, in the class of nested models that we estimated, REE may or may not be incorporated in the gain and loss subvalues, that should be thought of as averages - though not arithmetic but of an exponential moving type in view of equations (1) - see page 32 in [50]. The second and key term in equation (2) specifically captures the decision weight attached to REE, and it can itself be decomposed into the decision weights on Jackpots and on Black Swans. More precisely, the indicator function 1JP (o) equals one if the option exposes to Jackpots (namely when the chosen option is either A or V) and zero otherwise. Similarly, 1BS(o) equals one if the option exposes to Black Swans (namely options F and V) and zero otherwise. Therefore, γg and γl reflect the (possibly different) weights attached to Jackpots and Black Swans in the decision to choose one particular option rather than any other available.

Finally, the associated so-called action probabilities are given, for each option, by the Softmax function (a.k.a. the multinomial Logit model)πt(o)=eVt(o)/∑i=14eVt(i), each a number between zero and one.

In our model-fitting procedure (described in detail below), the key parameters of interest are the decision-weight terms for rare and extreme gains (γg) and losses (γl). In the baseline Q-learning model, both terms are fixed at zero, meaning that REE do not receive any special weighting beyond their contribution to standard value updates. We therefore compared two families of candidate models to this benchmark: one in which REE contribute directly to option sub-values, and another in which REE influence choice only through dedicated decision-weight parameters (γg,γl).

The latter class of models is especially informative. If selected, it suggests that animals treat REE as qualitatively distinct inputs to the decision process-processed separately from average reward estimates-rather than integrating them into conventional value updating. Overall, the augmented Q-learning framework includes up to eight free parameters: learning rates for gains and losses, forgetting rates for gains and losses, decision weights for gain- and loss-related sub-values, and the two REE-specific decision weights γg and γl.

Model comparison using information criteria (summarized by phenotype in Table 1) shows that 17 of 20 animals incorporate REE using dedicated decision weights, whereas only three animals are best fit by models in which REE carry no special computational status. Most rats (including 8 blue, 4 green, and the red and pink animals) rely specifically on γg and/or γl to shape choices, with or without distinct forgetting parameters. A minority of blue rats incorporate REE both through decision-weight parameters and through REE contributions to sub-values. Learning and forgetting rates did not systematically distinguish between model classes.

Figure 5 presents the predicted sensitivities (bottom left) and simulated sensitivities (bottom right), compared with empirically observed sensitivities (top; identical to Figure 2, right panel). Predicted sensitivities - computed using each rat’s estimated parameters and the actual reward sequence - closely reproduce empirical sensitivity patterns. Simulated sensitivities were obtained by running each selected model through artificial 41-session datasets, using median parameters for the blue and green phenotypes (and individual parameters for the red and pink rats). These simulations show that the selected models generate stable behavioral phenotypes for the blue, red, and pink animals. Green rats, however, tend to segregate into two stable subgroups along the vertical (One-Sided Sensitivity) axis.

Total Sensitivity to REE (y-axis) against One-sided Sensitivity to REE (x-axis).

Averages for each rat over the 41 sessions; top panel replicates the right panel in Figure 2: observed sensitivities; bottom left panel: predicted sensitivities derived from selected models; bottom right panel: simulated sensitivities computed from runs of selected models over artificial 41 sessions

Estimated parameters for each rat are shown in Figure 6. A prominent distinction between phenotypes emerges when focusing on the REE-specific d ecision w eights: b lue rats show a strong selective sensitivity to Black Swan, expressed as significantly n egative γl values (p = 0.0102). This indicates a robust computational penalty assigned to options that expose the animal to Black Swans (options F and V), consistent with their observed behavioral avoidance of these outcomes. In contrast, green rats do not show this selective aversive weighting and typically combine options F and A, suggesting a less selective treatment of REE. The before/after analysis illustrated in Figure 4 shows that the behavioral response to Black Swans is locally small in terms of both Total and One-sided sensitivities. This suggests that such effects are likely to be too subtle to be captured by this class of models for most rats.

Parameters of the selected Q-Learning model, estimated for each rat over the 41 sessions.

Left panel: decision weights for averages of gains and losses and for REE; right panel: learning and forgetting rates for gains and losses; while individual parameter for pink and red rats are represented as dots, parameters for blue and green rats are represented by box plots with 10th and 90th percentiles, interquartile ranges and median; between phenotype comparisons carried out by Wilcoxon Rank Sum tests (⋆ means p < 0.05) - see Material and Methods subsection Augmented Q-Learning Model Estimation and Simulation for details

Discussion

Asymmetric behavioral responses to rare extreme losses and rare extreme gains

We showed, first, that all rats in our sample exhibit some degree of Total Sensitivity to REE, with most animals displaying medium to high sensitivity. Second, we identified a dominant behavioral phenotype characterized by near-complete Black Swan Avoidance combined with partial Jackpot Seeking. We interpret these findings as reflecting a fundamental asymmetry between negative and positive REE, with a few caveats. Though we cannot fully exclude the possibility that satiety contributed to the lack of complete Jackpot Seeking, we did not systematically observe animals that stopped seeking the Jackpot. Because we restricted the rare outcome to occur on either the 10th or the 60th activation in a session (Table S1 in Supplementary Material), the question of whether the animals learn this association may arise. If the animals had learnt the 10th and 60th activation, they would exhibit a choice strategy that would tend to be more optimized than what is observed. For example, the options offering the possibility to obtain the Jackpot are not optimal in terms of gains for the frequent events, therefore the animals should tend to select these options only around the 10th and 60th choice. Most of their other choices should favor the options delivering the larger gains in the frequent domain. This is not what is observed.

In our task, avoiding Black Swans requires accepting longer frequent delays, and seeking Jackpots requires foregoing larger frequent gains. However, rats can guarantee the avoidance of Black Swans by selecting convex loss schedules, making these options relatively attractive. In contrast, no sequence of choices can guarantee the occurrence of a Jackpot-its realization depends on rare stochastic draws. For illustration, if 150 choices are made in a session (within the observed range) and the ex-ante REE probability is 1%, a Poisson approximation gives a probability of ≈ 78% for observing exactly one REE-a non-negligible chance for zero occurrence. Thus, our design captures a general asymmetry likely to characterize positive versus negative REE more broadly: while extreme losses can in principle be fully avoided, extreme gains cannot be assured, even under persistent optimal seeking. This asymmetry has important consequences for choice behavior. Rats experience a continuous stream of evidence-derived from the frequent-outcome domain-that points toward concave options as locally optimal. In particular, the Fragile option yields higher frequent gains and smaller frequent losses, and stochastically dominates the other options when REE do not occur. Although Fragile is dominated by Anti-fragile when both the Black Swan and Jackpot occur, complete Black Swan Avoidance provides near-certain protection against extreme losses. This makes strategies combining Anti-fragile and Robust options attractive, despite violating Stochastic Dominance in the frequent domain: full-domain Stochastic Dominance is effectively secured.

In contrast, complete Jackpot Seeking is less attractive because it cannot ensure the occurrence of a Jackpot; the cost of receiving smaller gains on most trials becomes unjustified unless the compensatory extreme reward actually occurs. Partial Jackpot Seeking therefore emerges as a more balanced strategy: it reduces violations of Stochastic Dominance in the frequent domain while still allowing some exposure to large rare gains. Put simply, bearing frequent relatively large losses is acceptable only if the avoided Black Swan is certain, and bearing frequent small gains is acceptable only if the compensating Jackpot is likely to appear-which it is not. Our results indicate that the “blue” rats have learned this property, combining Robust and Anti-fragile options to achieve near-complete Black Swan Avoidance with partial Jackpot Seeking. In contrast, “green” rats continue to mix Anti-fragile and Fragile options, leaving them exposed to Black Swans.

Interestingly, since rats do not stick to one choice but show some flexibility in their strategy, this rules out the influence of a spatial bias as the main driver in their response. Interestingly indeed, the design of the task was biased in two ways: the favorite hole during training was allocated to the antifragile option as a proof of concept and the REE occurred at the 10th and 60th choice if the appropriate option was chosen. Since most rats diversified their choices, the biased location does not seem to have prevented rats to learn all options. One might wonder whether or not they may have understood and learnt the 10th and 60th choice REE delivery. If this was the case, a trend towards optimal choice such as options delivering the largest rewards and the lowest losses in the most frequent domain with a shift towards options with positive REE and avoidance of negative REE around the 10th and 60th trial should have been observed. This was never the case.

In theory, because frequent outcomes occur often, animals should be able to detect Stochastic Dominance in the frequent domain relatively quickly. However, prior work indicates that animals require many sessions to learn Stochastic Dominance (e.g. [59], [3]). Our findings suggest a possible explanation: substantial experience may be needed to infer that REE have not occurred and can be excluded from the estimated distribution.

At this stage, it is natural to ask whether our findings reflect loss aversion or ambiguity aversion. Although most studies do not focus on REE, tasks involving both positive and negative outcomes generally report loss aversion (e.g., [18], [14], [61]), consistent with Prospect Theory ([22]). Models based on this framework, however, tend to fit rat behavior more effectively in tasks involving o nly rewards, w here losses correspond to the a bsence of a positive outcome rather than punishment ([10],[3]). Emotional and anxiety-related mechanisms may also contribute ([60]), potentially implicating the basolateral amygdala, which has been linked to differential evaluation of negative outcomes in rats ([54]). Ambiguity aversion in non-human animals remains relatively unexplored (but see [44]).

Our results tentatively suggest that rats may exhibit ambiguity reduction with respect to REE: combining near-complete Black Swan Avoidance with only partial Jackpot Seeking collapses the sampling probability of Black Swans toward zero, while keeping the sampling probability of Jackpots bounded between zero and ε. This behavioral asymmetry provides further evidence that extreme gains and losses receive distinct treatment during decision making. Classical Q-learning models, which allow differential weighting of gains and losses, fail to capture these patterns. In contrast, the augmented Q-learning models incorporating explicit sensitivity to REE accurately reproduce the observed behaviors - principally through the strong aversion to Black Swans that distinguishes the “blue” phenotype from the “green” one.

While a wavy recency effect has been documented in human experiments ([38]), the before/after analysis reported in Figure 4 suggests that there is no sizeable immediate recency effect for Jackpots in rats. Even for Black Swans, the immediate recency effect we report remains modest when using a 10-trial window, and the analysis of the choice immediately following a REE does not show evidence of immediate negative recency. Although further investigation is beyond the scope of this paper, possibly interacting recency effects of Jackpots and Black Swans seems a promising research avenue. Also note that the task design does not allow rats to completely avoid negative rare events (RE) unless they cease performing the task altogether - a pattern typically seen in paradigms involving aversive stimuli such as electric foot shocks. The fact that all 20 rats maintained stable performance across the 41 sessions therefore provides evidence against a pronounced “hot-stove effect” (see [12]).

The augmented Q-learning models suggest a specific neural pathway for REE encoding

The close fit and predictive accuracy of the augmented Q-learning model for the dominant behavioral phenotype point to a novel neurobiological hypothesis. Within this modeling framework, 11 of the 13 rats in the blue group (as defined in the Materials and Methods) appear to exclude REE from the Q-value update, yet display a specific sensitivity to Black Swans through a negative parameter γl. In other words, when evaluating available actions, “blue” rats do not integrate Black Swans into the running average of losses; instead, they assign them a distinct decision weight, effectively reducing the subjective value of options that expose the animal to these extreme losses.

In humans, rare and extreme stimuli are known to be disproportionately salient in perception and memory compared to moderate ones (e.g., [26]). Our behavioral and modeling results suggest that rats may exhibit a similar cognitive bias. This, in turn, raises the possibility that REE - particularly rare and extreme losses-may rely on neural encoding mechanisms that differ from those that process frequent, small rewards, which are classically attributed to dopamine neurons ([9], [46], [61], [55]), the striatum ([7], [11]), and cortical regions such as the anterior cingulate and orbitofrontal cortex ([45], [15]).

Identifying where and how REE are encoded in the brain remains an open question. If integrating REE requires representing long-term or low-frequency outcomes-beyond the short-term reward horizons typically studied (e.g., [56]) - candidate regions may include the anterior insula and the amygdala. The latter, given its central role in emotional salience, is a plausible substrate for representing highly aversive and unpredictable outcomes, which are known to bias choice asymmetrically toward negative events ([7], [54]).

From a computational standpoint, it is statistically rational to avoid folding extreme outliers into the mean of a distribution, as this can destabilize the estimate. Analogously, adaptive decision-making may benefit from treating REE separately from frequent outcomes, and from imposing different weights on Black Swans versus Jackpots. If such differential encoding confers evolutionary advantages in volatile environments (see [8]), then we might expect dedicated neural mechanisms to have evolved-possibly across multiple species.

To our knowledge, this specific hypothesis, grounded jointly in behavioral data and computational modeling, has not yet been directly tested in neuroscience. Doing so would align with the computational-validity framework advocated in [42], and represents an important direction for future research.

Relevance for humans

Whether humans would also find Black Swan Avoidance a nd Jackpot Seeking attractive - and how they might combine these strategies in a similar experimental context-is an obvious question raised by our results. Our experimental design can be readily adapted to human participants, allowing us to test whether people likewise tend to avoid harmful REE and pursue beneficial ones. We have started exploring modified designs for human subjects in which the rare but non-extreme outcomes are removed, in line with [27] who emphasize that combining rarity and extremity is key. Preliminary results indicate that the behavioral phenotypes observed in rats also emerge in humans under these modified conditions, suggesting that REE are the primary drivers of observed behaviors. Furthermore, such an extended task may be a promising tool to delineate decision-making profiles in professional environments (e.g. firemen) and also to investigate vulnerability to addiction in patients. This translational potential makes the present findings i n r ats r elevant t o b roader issues concerning the past and future of humanity.

Within the framework of the Anthropocene, mounting evidence shows that climate change and other environmental disruptions driven by human activities generate extreme events (see e.g., [13]), including rare but potentially catastrophic ones that could threaten many species with partial or complete extinction. Although uncertainty remains about the magnitude of these risks, empirical data place humanity in a situation that raises difficult but pressing questions. Why has our species historically failed to avoid behaviors and policies that increase exposure to destructive REE? And to what extent will humans be able to cope with such events in the near future, should they occur? Put differently, are humans following trajectories that diverge from Black Swan Avoidance rather than moving toward it?

While it is intuitive to assume that avoidance of destructive REE-and the ability to cope with them when they occur - provides adaptive value, compelling evidence for such evolutionary mechanisms remains scarce (but see [8]). If correct, this idea introduces a paradox: humans disproportionately contribute to creating environments rich in harmful REE. Does our species possess a particular trait that impairs our ability to protect ourselves against them, despite the apparent evolutionary benefits of doing s o? Because destructive REE are increasingly endogenously generated by human activity, one may even wonder whether a uniquely human developmental or epigenetic factor predisposes us to decisions that fail to eliminate exposure to extremely harmful outcomes. If so, such a factor could also impair our ability to cope with REE once they become unavoidable. For instance, humans might struggle to detect predictive cues of REE - such as convexity in our experimental design (see [2]).

The results from our rat experiments fortunately do not support this pessimistic view. The substantial individual variability we observe-where some animals fail to avoid rare but extreme losses even when given the opportunity - suggests that humans might share this variability rather than being uniquely deficient. Importantly, the majority of rats consistently chose options that nearly eliminated exposure to harmful REE while still allowing partial access to beneficial REE in the form of Jackpots. This “adaptive” subgroup constitutes the largest portion of our sample.

A speculative hypothesis follows: some humans, and perhaps many, may also adopt strategies that avoid harmful - and sometimes pursue beneficial - rare extreme outcomes. More broadly, if sensitivity and behavioral responses to REE vary across individuals within a species, this raises important questions about how such heterogeneity shapes population dynamics, social structures, and, crucially, collective decision-making.

Material and Methods

Experimental Model and Subject Details

Animal Subjects

Adult Lister Hooded males (n = 20, ≈200 g at arrival, Charles River) were housed in groups of two in Plexiglas cages and maintained on an inverted 12 h light/dark cycle (light onset at 7 pm) with water available ad libitum, in a temperature - and humidity - controlled environment. Food was slightly restricted (≈80% of daily intake). Animal care and use conformed to the French regulation (Decree 2010-118) and were approved by local ethic committee and the French Ministry of Agriculture under #03129.01.

Experimental Method Details

Apparatus

All behavioral experiments took place during the animals’ dark phase in standard five-hole operant boxes (MedAssociates) located in ventilated sound-attenuating cubicles. One side of each box was equipped with a central house light, a tone generator and a food magazine, outfitted an infrared beam for detecting nose poke inputs. Sucrose pellets (20 mg; Bio-Concept Scientific) were delivered from an external food pellet dispenser. An array of five response holes was located on the opposite curved wall, each equipped with stimulus lights and infrared beams for detecting input (nose poke). The center hole was continuously closed throughout the experiments (Fig. 1a). Data were acquired on a PC running MedPC-IV.

Design of Behavioral Sequences

Four menus were elaborated by mixing convex and concave exposures for both the gains (sugar pellets) and the losses (time-out punishment) described on Figure 1 (panel (b)). For the gains, animals could obtain 1, 3 (NE domain, blue), 12 (RE domain, yellow) or 80 (REE domain; pink) sugar pellets in the convex exposure or 2, 4, 5 or 5 pellets in the concave exposure. For the losses, convex exposure may impose 6, 12, 15 or 15 sec of time-out punishment while it was 3, 9 (NE domain; blue), 36 (RE domain, yellow) or 240 sec (REE; pink) for the concave exposure.

The four behavioral options depicted in Figure 1, panel (c), are therefore combinations of the above convex and concave exposures: the “Anti-fragile” exposure at the top middle is convex in both gains and losses, while the “Robust” option (left) is only convex for the losses. On the other hand, the “Vulnerable” exposure (right) is convex only for the gains, while the “Fragile” option (bottom middle) is concave for both gains and losses. This implies that the extreme - but rare - gain of 80 pellets (i.e. Jackpot) may be delivered only when either the Anti-fragile or the Vulnerable options are picked, while the extreme and rare time-out punishment of 240 sec (i.e. Black Swan) may be only experienced if choosing either the Fragile or the Vulnerable options.

The first three events of each behavioral options belong to the frequent domain as their frequency of occurrence, respectively 0.5, 0.4 or 0.1, is significantly larger than zero. During behavioral testing, animals equivalently experienced the gain and loss domains. On the other hand, extreme outcomes, i.e. Jackpot or Black Swan, have a much smaller likelihood of occurring since they may appear only at particular point during the behavioral sequences (see below). This means that they could happen if a rat has chosen an exposure that is either convex in the gain domain or concave in the loss domain at a given time. This implies that the frequency of rare and extreme events is less than half of one percent for all rats. Despite their low probability, extreme events have some importance because of their value. Our calibration of both frequencies and outcomes ensures that concave exposures dominate convex exposures: if the extreme gain (Jackpot) does not materialize, the expected payoff in sugar pellets is larger for concave exposures than convex exposures (i.e. ex-post first order stochastic dominance, over the frequent domain). Similarly, the expected time-out punishment is lower for concave exposures compared to convex ones, if the Black Swan does not happen. However, ex-post first-order stochastic dominance is reversed in the presence of extreme events, in which case convex exposures become more interesting in terms of payoff than concave ones. The reason we imposed such a dominance reversal was as follows. If rats always choose concave exposures, then they show no sensitivity to rare and extreme events, since they act as if those events never occur and always go for first-order stochastic dominance. This gives us our first measure, Total Sensitivity to REE, that simply sums up the proportion of convex exposures that are chosen for each rat over the 41 sessions. Because we deliberately integrate gains and losses in our design, we need to distinguish whether rats tend to choose convex exposures symmetrically over the gain and loss domains. We say that a particular rat exhibits Black Swan avoidance when it picks convex exposures in the loss domain more often than in the gain domain. Likewise, Jackpot Seeking occurs when convex choices are more frequent in the gain domain.

To avoid potential learning of the event occurrence during behavioral training and testing, ten different sequences of events, with respect of the first-order stochastic dominance as well as the balance between gain and loss domain exposures, were generated and randomly used for behavioral training and testing (Supplemental Table S3). Furthermore, to increase the rarity and the unpredictable nature of extreme events, the sequences of events used during behavioral training and testing were declined into seven various sequences, in which Jackpot and Black Swan are either unavailable, solely or both available at a given time point of the sequence of events (Supplemental Table S1): when available, extreme events could be obtained at the 10th or the 60th activation within a given sequence, but could not occur at the same time. For example, in a behavioral sequence where the Jackpot should be available at the 10th position, any poke in the Anti-fragile and Vulnerable holes following nine responses made in one of these menus (regardless of the positive or negative outcomes) would trigger the delivery of the Jackpot. Of note, depending on the sequence used, animals could experience both extreme events in a single session.

Behavioral Training and Testing

Training was divided into five distinct phases before the final test: acquisition of the food collecting responses, acquisition of nose poking in the holes, training with four holes, attribution of menus to hole and training on the menus (Figure 7). Each session started with the illumination of the house light.

Time-line of experiments.

Acquisition of the food collecting responses: Animals were trained to collect sucrose pellets in the food magazine during three 30-min daily sessions (100 pellets max) under fixed ratio 1 schedule of reinforcement (FR1): one nose-poke into the food magazine triggered the delivery of one sucrose pellet. During this initial phase, nose-poke in any hole had no programmed consequences.

Acquisition of nose-poking in the holes: Here, animals had only access to one hole (the three others being occluded) during the 20-min sessions. One nose-poke in the available hole triggered the illumination of the hole-light and the delivery of one sucrose pellet in the food magazine. Perseverative nose-pokes (those performed before food collection) had no consequences. Following food collection in the food tray, animals were allowed to poke again in the opened hole. During each session, a maximum of 100 pellets were delivered. All animals were trained twice on each hole.

Training with four holes: Following the eight training sessions, animals were allowed to poke in the four different holes during twenty daily sessions of 20-min. Nose poke in any hole triggered the illumination of the associated light and the delivery of one sugar pellet in the food magazine. Both perseverative activations and pokes in other holes had no consequences. After food collection in the magazine, animals were allowed to nose poke again in any hole. The first ten training sessions were limited to 100 sugar pellets. During the following ten sessions, animals were able to collect up to 200 pellets per session.

Attribution of menus to hole: We determined the spatial preference for each rat by establishing the percentage of activation of each hole during the last ten sessions. To favor the emergence of Anti-fragile choices, menus’ attribution was made as follow:

Anti-fragile exposure was associated to the preferred hole

Robust exposure was associated to the 2nd preferred hole

Vulnerable exposure was associated to the 3rd preferred hole

Fragile exposure was associated to the least preferred hole

Training on the exposures: Here, animals were first trained on the gain domain, i.e. no time-out punishment, for each menu/hole. They were subjected to two 20-min sessions, with unlimited number of pellets, during which only one hole was available (eight training sessions in total). For Anti-fragile and Vulnerable options, we used the two sequence-types where the jackpot was available at the 10th and 60th activations (Supplemental Table S1) to ensure that all individuals could experience both an early and delayed jackpot during training. Animals were then allowed to explore all gain options (four opened holes) during 9 20-min sessions, for which different behavioral sequence-types were used. Before training on the loss domain of the different menus, animals were first exposed to a mild and constant 3-sec time-out punishment. Here, animals had access to all options, but half of the activations lead to a 3-sec time-out punishment, notified by a 3-sec tone and the extinction of the house light. During this period, pokes in the different holes or the food magazine had no programmed consequences. After the 3-sec time-out punishment, the house light was turned on and the animals could again pokes in the different holes. Following nine 20-min training sessions, the loss domain of each menus was progressively introduced. As described above, each time-out punishment was notified by a 3-sec tone and the extinction of the house light for the whole duration of the punishment, which termination was signaled by house light illumination. Animals were first exposed to concave exposures, having only access to vulnerable and fragile holes during four 20-min sessions. They were then exposed to convex losses (only Anti-fragile and Robust holes available) for another four 20-min sessions. Thus, at the end of the training, all animals experienced both extreme events at least four times.

Final tests: For the final tests, animals were free to explore all menus throughout the forty 20-min sessions. Animals experienced four times the ten different sequences, randomly distributed across sessions. Population (n = 20) was subdivided into two groups that experienced different, but equivalent, sequence-type distribution (Supplemental Table S2).

Figure 7 depicts a simplified time-line of the experimental procedure.

Table S3 shows the chain of events in the ten sequences used for behavioral training and testing. Numbers in the second column indicate which event would occur and the sign preceding it whether it belongs to the gain (no sign) or loss (minus sign) domains. For example, the third event in the first sequence (noted −2) triggers a time-out punishment of 9 or 12 sec, depending on whether an animal performed his third nose poke in concave or convex menu, respectively. Note that extreme events (which should be noted 4 or −4) do not appear in the sequences, as they would automatically replace the 10th or 60th events of the sequence. See supplementary material for detailed information on sequences of events.

Modelling and Statistical Analysis

Modelling Convex and Concave Exposures

Central to our experimental design is the notion of convex/concave exposure under radical uncertainty, that is, when probabilities and consequences are unknown a priori to subjects. A well known measure of convexity is the Jensen gap that is derived from Jensen’s inequality (see [30] for a graphical exposition), which we now define and relate to statistical moments. In our context, the relevant form of Jensen’s equality states, loosely speaking, that the expectation of a convex function of a random variable is larger than the value of that function when evaluated at the expectation of the random variable. Jensen’s gap is then defined as the difference between the expectation of the function minus the value of the function at the expectation (hence positive by construction). The inequality is reversed for a concave function and the (positive again) Jensen’s gap is then defined as the difference between the value of the function at the expectation minus the expectation of the function. In the gain domain, rats obtain 1, 3, 12 or 80 sugar pellets if the convex exposure is chosen, or 2, 4, 5, 5 pellets if the concave exposure chosen. In the loss domain, convex exposure imposes 6, 12, 15 or 15 seconds of time-out punishment, and 3, 9, 36, 240 seconds for the concave exposure. Because the values in the loss domain are proportional to the values in the gain domain, the former corresponding to 3 times the latter, we focus here on values for pellets. The statistical properties of convex and concave losses follow accordingly. More formally, sequences of gains are ordered by increasing values and are denoted {xiX}i=14={1,3,12,80} for the convex exposure and {xiV}i=14={2,4,5,5} for concave gains. We assume identical probabilities {pi}i=14={p1,p2,p3,ε} for both concave and convex gains, where ε is the ex-ante probability of the rare and extreme event (REE for short). The third value (that is, 12 for convex gains and 5 for concave gains) is labeled a rare event (RE), and the sets of the lowest two values (that is, 1 and 3 for convex gains, 2 and 4 for concave gains) for both exposures are composed of normal events (NE).

Making use of the exponential transform, we define Jensen gaps for the convex and concave exposures as, respectively, JGX=∑i=14piexiX−e∑i=14pixiX,JGV=∑i=14piexiV−e∑i=14pixiV which relate to statistical moments as follows. The moment generating function for the convex and concave exposures are given by, respectively: MX(t)=∑i=14pietxiX and MV(t)=∑i=14pietxiV. The series expansion of the exponential function allows us to derive: MX(1)=1+m1X+12!m2X+13!m3X+⋯+1N!mNX+…MV(1)=1+m1V+12!m2V+13!m3V+⋯+1N!mNV+… where mNX and mNV are the N-th statistical raw moments of the convex and concave exposures, respectively. For example,mNX=∑i=14pi(xiX)N for any integer N ≥ 1. Jensen’s gaps for both exposures are then given by JX=MX(1)−em1X,JGV=MV(1)−em1V. Using again the series expansion of the exponential of the first moment for each exposure, it follows that both positive Jensen gaps can be written in terms of moments, as follows: (3) JGX=∑k=2∞mkX−(m1X)kk!,JGV=∑k=2∞mkV−(m1V)kk! Note that the sums in equation (3) nicely allow for a straightforward decomposition in terms on all raw moments with order larger than two.

More specifically, we set probabilities to {pi}i=14={0.5−ε/3,0.4−ε/3,0.1−ε/3,ε} identically for both concave and convex gains. Treating the probability for the REE as a varying parameter, it can be shown that Jensen’s gaps for convex and concave exposures are ordered, with the former larger than the latter when ε ≤ 0.02. Over that range, it also holds true that both Jensen’s gaps are monotone increasing function of ε. While this property also holds for the expectations and variances of both exposures, it doesn’t for higher-order moments. For example, the skewness and kurtosis of the convex exposure turn out to be hump-shaped functions of ε, with peaks corresponding to values of ε smaller than one percent. This fact underlines that convexity measured by Jensen’s gap offers a unifying approach to rank exposures, seen as lotteries, whereas statistical moments do not necessarily do so. We conjecture that, more generally, all lotteries satisfying our assumptions (including monotone probability distribution) can be ranked according to their convexity as measured by the Jensen gap.

In Table 2, we report the central moments and the log of Jensen gaps in the top five rows, for both exposures when ε = 1/75 ≈ 1.3%, which corresponds to obtaining one REE out of 75 nose pokes. For that particular parametrization, convex and concave exposures differ mostly in their respective kurtosis and in their Jensen gaps. From row six and below, we report the decomposition of Jensen gaps in terms of the raw moments. For example, in row six is given the ratio of (m2 − (m1)2)/2 - that is, half the variance - to the corresponding Jensen gap for each exposure, in percentage. Strikingly, while moments of order 2 to 8 explain about 90% of the concave exposure’s Jensen gap, they contribute negligibly to that of the convex exposure. In fact, it turns out that moments with order around 80 start contributing, although for a small share each, to the convex exposure’s Jensen gap. In sum, while for the concave exposure a few low-order moments concentrate the contributions to the Jensen gap, the contributions of raw moments to the convex exposure’s Jensen gap spread across a larger number of very high order raw moments.

We derive next the ex-ante properties of the probability distributions associated with concave and convex gains (for definitions of stochastic dominance and congruent utility classes, see for example Fishburn and Vickson [16]; also note that first-order, resp. second-order, stochastic dominance implies second-order, resp. third-order, stochastic dominance):

In the domain restricted to NE, concave gains first-order stochastically dominate convex gains. In addition, concave gains then have a larger expected value than that of convex gains, with equal variance, skewness and kurtosis for concave and convex exposures.

In the domain restricted to NE and RE, concave gains second-order stochastically dominate convex gains. In addition, concave gains then have a larger expected value and smaller variance, skewness and kurtosis than that of convex gains.

In the full domain including NE, RE and REE, convex gains second-order stochastically dominate concave gains if and only if ε ≥ 0.302%. In addition, convex gains then have a larger expected value, variance, skewness and kurtosis than that of concave gains if ε ≥ 0.302%.

The above assumptions on the experimental design are stated in terms of stochastic dominance and moments of the probability distributions. They also have implications in relation to standard approaches to decision-making. First, from an ex-ante perspective with perfect information about the probability distributions of all exposures, value-maximizing subjects whose preferences are represented by any non-decreasing value function would choose concave gains in the domain restricted to NE. They would continue to do so in the domain restricted to NE and RE for any non-decreasing and concave value function. However, in the full domain with REE, value-maximizing subjects endowed with any non-decreasing and concave value function are predicted to choose convex gains.

We denote V({xi}i=1n,{pi}i=1n) the value attributed to sequences of gains {xi}i=1n with probabilities {pi}i=1n, for n ≤ 4. The above assumptions have the following implications in terms of value maximization:

In the domain restricted to NE, V({xiV}i=12,{pi}i=12)>V({xiX}i=12,{pi}i=12) holds.

In the domain restricted to NE and RE, V({xiV}i=13,{pi}i=13)>V({xiX}i=13,{pi}i=13) holds.

In the unrestricted domain, V({xiX}i=14,{pi}i=14)>V({xiV}i=14,{pi}i=14) holds if and only if ε ≥ 0.302%.

Note that the reversal in second-order stochastic dominance (and value) in the full domain favors convex gains when extreme events that are indeed very rare - with probabilities much smaller that one percent - are added. In contrast, absent REE, adding only a RE is not enough to favor convex gains, even though it allows subjects to possibly detect convexity/acceleration and concavity/deceleration of gains and losses.

The above assumptions hold true under expected utility, that is when V(⋅,⋅)=∑i=1npiu(xi) with the appropriate conditions on the utility function u (non-decreasing for first-order stochastic dominance, non-decreasing and concave for second-order stochastic dominance; see [16]). To better match experimental data, however, expected utility is increasingly supplemented by some form of probability weighting w(pi). For instance, [10] use a two-parameter functional form due to [39] and find that 30 rats out of 36 behave as if their probability weighting w(pi) is concave. In that case, the above ex-ante properties hold mutatis mutandis, for example, under expected utility with probability weighting, defined as ∑i=1nw(pi)u(xi). Alternatively, in the setting of rank-dependent expected utility, we can make use of results 3 and 4 in [25] to show that the above ex-ante properties still hold provided that the transformation of the cumulative distribution function is any increasing-concave function. Note that since in our experimental design REEs are both extremes (in the sense of being the largest values) and rare (that is, they have very low probabilities), one expects similar results under probability weighting and rank-dependent expected utility (or cumulative prospect theory for that matter). Although in theory assuming simply a weighting function w(pi) that is inverse-s-shaped and “very” convex for large gains (see [39] for the associated parametric restrictions) could overturn the rankings of concave and convex exposures stated above, results in [10] suggest that this is not to be expected for the overwhelming majority of the rats that are subject to their experiments.

Choice Data Analysis

Given the four options modelled in the previous section, each of the 20 rats went through 41 final sessions. Denote fA, fF, fR, and fV the relative frequencies (in terms of nose pokes) with which each of the four options Antifragile, Fragile, Robust and Vulnerable, respectively, is chosen over a particular session. In order to represent the rats’ choices, we construct the rotated square in panel (d) of Figure 1 using a linear transformation of the 3-frequency vector, as follows: (4) (TSREEOSREE1−fF)=(2110−11111)(fAfRfV) In equation (4), the first two rows deliver the two coordinates of interest that are the Total Sentivity to REE (TSREE in short) and One-sided Sentivity to REE (OSREE), while the last row requires that all frequencies sum to one. To derive TSREE and OSREE, it is useful to go through the following steps. Since we are interested in the convexity property of each chosen option, we first measure the frequencies of convex options in the gain domain by summing fA and fV. Similarly, fA + fR measures how frequent choices are convex in the loss domain. It follows that TSREE measures total convexity if it equals 2fA + fV + fR, while OSREE equals fV −fR and measures asymmetric convexity (that in the gain domain minus that in the loss domain). Hence the first two rows in equation (4). Using that fA + fF + fR + fV = 1 - that is, the last row of equation (4) - it follows that TSREE equals the simpler expression 1 + fA − fF. Note that the 3 × 3 matrix in equation (4) is invertible, which implies that all three frequencies can be recovered from the left vector. In addition, that matrix is not unique in the sense that the definitions of TSREE and OSREE that appear above as its first two lines could be moved to any other rows, provided that the identity in the third row and the constraint in the last row are adapted accordingly. This essentially means that given both differences TSREE (= 1 + fA − fF) and OSREE (= fV − fR), any of the four relative frequencies provides enough information to derive the remaining three, using the constraint that all four frequencies sum up to one. Finally, it follows that in the rotated square in panel (d) of Figure 1, the four vertices have the following coordinates: (0, 0) for Fragile, (0, 2) for Antifragile, (−1, 1) for Robust, and (1, 1) for Vulnerable. TSREE and OSREE are depicted in Figure 2.

TSREE and OSREE are also related to behavioral measures of the share of nosepokes that lead to either exposure or non exposure to REE, as follows: we define Black Swan Avoidance as the share of nosepokes that lead to avoidance of Black Swans - i.e. fA + fR - and Jackpot Seeking as the share of nosepokes that lead to be exposed to Jackpots - i.e. fA + fV. It follows that TSREE is by definition the sum of those two shares, while OSREE is the difference between the former and the latter. Formally, if we denote JPS and BSA the shares of nosepokes that lead to be exposed to Jackpots and to avoid Black Swans, this implies that JPS= 0.5(TSREE+OSREE) and BSA= 0.5(TSREE−OSREE). In other words, in the (not rotated) square that appears in panel (d) of Figure 1, the four vertices have the alternative coordinates: (0, 0) for Fragile, (1, 1) for Antifragile, (0, 1) for Robust, and (1, 0) for Vulnerable, in the (JPS,BSA) coordinates. Obviously, both interpretations are equivalent up to a change in coordinates which is a bijection. Note that in the rotated square in panel (d) of Figure 1, lines with a 45-degree slope depict choices with constant BSA so that lines moving north-west represent increasing BSA. Similarly, lines with a −45-degree slope depict choices with constant JPS so that lines moving north-east represent increasing JPS. Median BSA and JP are depicted in Figure 3.

In Figure 4 we report the short term behavioral responsiveness of rat to REE. Behavioral responsiveness is measured by calculating for each rat before/after differences in TSREE and OSREE. This amounts to calculate TSREE and OSREE over the 10 choices preceding a REE and over the 10 choices following the occurrence of each REE. The window of 10 choices before and after is dictated by the configuration of REE in our d esign. A REE can happen at 11th choice in a sequence (see appendix). Before/after TSREE and OSREE are averaged by rat over the 41 sessions in the gain domain (Jackpot) and in the loss domain (Black Swan). Table 3 reports the values for the mean before/after differences. I n each panel of Figure 4, each rat is depicted by a color-coded dot and the dotted black line represents the 45-degree line. Color-coded dot on the 45-degree line indicate no difference in TSREE or OSREE before and after a REE, i.e. no short term behavioral responsiveness to REE. To visualisation purpose, we computed for each panel a smooth spline regression that estimates a non parametric relationship between before and after TSREE and OSREE. They are plotted as solid black lines in each panel.

We then test the statistical significance of short term behavioral by conducting bootstrap paired-sample mean tests on before/after coordinates (separately) for each type of REE. We assessed short-term behavioral responses using a paired bootstrap test (boot.paired.bca) in R (wBoot package). This non-parametric method estimates the mean difference between paired conditions by resampling with replacement and computes a bias-corrected and accelerated (BCa) confidence interval, providing robust inference without assuming normality. An observation in the sample is the mean behavioral responsiveness in TSREE or OSREE for a given rat. Observations therefore respect statistical independence. Bootstrap tests lead to the following p-values, under the null hypothesis that the mean difference i s zero. Following Jackpots, p = 0.3470 for TSREE and p = 0.0522 for OSREE, which indicates that the null hypothesis is not rejected at 1%. Following Black Swans, however, the null hypothesis is rejected at 1% since p = 0.0045 for TSREE and p = 0.0064 for OSREE. In sum, the mean before/after difference following a REE in the loss domain is s ignificant for both Total and One-sided sensitivities, further confirming the Black Swan avoidance result that we document in this paper.

We also report in Figure 8 how choices by rats after the 41 final sessions compare to choices made in the training step 3, defined in the section on Behavioral Training and Testing. We do so by computing TSREE and OSREE in step 3 sessions and pair coordinates resulting from training with that resulting from the 40 experimental sessions reported in Figure 2. The lines between two color-coded dot connects for each rat its conditioning coordinates with its coordinates resulting from its behavior in the 41 sessions. We report on the right side of the figure the total variation distance between the distribution of nosepokes in the 4 holes in the training and the distribution in the 41 experimental sessions. The interpretation of total variation distance in our setting is simple: it measures, for each rat, the proportion of nosepokes that need to be changed in order to equalize the behavior in the training and the behavior in the 41 final experimental sessions. Median and mean total variation distance are 0.293 and 0.324 respectively, which implies that half of the rats changed more than 30% of their choices in the final experimental sessions compared to the training sessions.

Variation of choices observed in final experimental sessions compared to that observed in training sessions (step 3 in Material and Methods section on Experimental Method Details).

Panel (a) shows variations of Total Sensitivity to REE (y-axis) and of One-sided Sensitivity to REE (x-axis), that is, dark color-coded dots replicate averages over the 40 sessions for each of the 20 rats - see panel (c) in Figure 2 - and are paired with averages over training sessions for the same rat, depicted with light color-coded dots connected through lines with corresponding dark color-coded dots; panel (b) shows total variation distance, that is, the proportion of total nose pokes in training sessions that need to be changed to replicate nosepokes in final sessions

Finally, in Figure S1 of the Supplementary Material, we report what we label “convexity premiums”, which are counterfactual situations that we construct as follows. Given the rats’ choices over the 41 sessions, we recall all random sequences that have been used to generate gains and losses for each rat. We next compute the outcome that would have been obtained, had the rat chosen convex options in the gain domain (that is, Antifragile or Vulnerable) and in the loss domain (that is, Antifragile or Robust) all the time. The right and left panels in Figure S1 depict the normalized convexity premiums thus computed, for each rat represented in rows, with convex exposures yielding more pellets and implying less waiting in terms of seconds.

Augmented Q-Learning Model Estimation and Simulation

Parameter optimization is carried out for each rat over the 41 sessions, using observed choices and outcomes. We estimate by maximum likelihood two series of augmented Q-Learning models. The first series integrates REE in Q subvalues whereas the second series does not. Each series consists of a first baseline model without specific forgetting rates and decision weights attached to REE. Two partial models integrate either specific forgetting parameters or decisions weights on the presence of REE. The fourth model integrate both. In total, we estimate eight nested models for each rat. Formally, we estimate four different models: a baseline model (Model 1) without specific forgetting parameters (αkf=αk, with k ∈ {g, l}) and with zero decision weights of REE (γg = γl = 0). Model 2 introduces specific forgetting parameters (αkf≠αk, with k ∈ {g, l}) and Model 3 introduces decision weights of REE (γg and γl≠0). Model 4 introduces both.

To ensure convergence and avoid local maxima, each parameter in baseline models was initialized with two different points of its parameter space. This makes in total 42 = 16 combinations of initial parameter values. Each of the 16 models is estimated using an automatized two-step procedure. The fist step consists of an unconstrained downhill simplex method in [31], which does not make use of first derivatives. Estimated parameters are then used as initial parameters in the second step of the estimation procedure, a quasi-Newton method in [17]. Parameter estimates from baseline models are then used as initial parameters for subsequent augmented models. Estimation is carried out using the above two-step procedure to ensure that we convergence is reached.

Once the eight models are estimated, we then compare them using both Akaike information criterion (AIC) and Bayesien Information criterion (BIC), that penalizes the use of additional parameters, so as to select which model configuration best fits observed behavior for each rat (see Supplementary Material for detailed results per rat). We check whether the selected model properly estimates sensitivities to REE by comparing estimated sensitivities to observed sensitivities to REE. This is done by computing Total Sensitivity to REE and One-sided Sensitivity to REE based on estimated choice frequencies from the extended Q-Learning model over the 41 sessions. Fitted sensitivities are reported bottom right panel in Figure 5. Finally, we checked that selected models were able to reproduce experimental data by running simulations for each phenotype observed in the sample of animals (as in [35]). Simulations are conducted over the artificially generated original 41 sessions, using individual parameter estimates for pink and red rats, median parameter estimates for blue and green rats and estimated Q subvalues as prior Qs. For each simulation, we then computed Total Sensitivity to REE and One-sided Sensitivity to REE. Simulations are reported in bottom right panel of Figure 5. See Supplementary Material for more details on the estimation results and for an exploration in related outcome range-adaption models